---
title: "Coffee Field Model: Affordances and a Process in Time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Coffee Field Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4
)
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
```

## What is a functional field?

A **functional field** is a model of how situational forces propagate
through an environment to produce action outcomes. At each moment the
field specifies which forces are *afforded* — which causal pathways are
open given the current state of the situation. A later step only matters
once an earlier step has changed the situation: this is **causal
gating**, and it is what makes a four-step plan into a connected causal
chain rather than four independent intentions.

The same field `F` supports two complementary workflows. **Induction**
runs data → `F`: you fit the field's coefficients from observed
behaviour with `pathXMY()` and friends. **Deduction** runs `F` →
behaviour: you treat the coefficients as a working theory and *simulate*
forward to see what a plan produces. This vignette is about deduction.
For the notation it uses throughout — the `F` / `F_t` field objects and
the `fZ_X.Y` parameter schema — see
`vignette("notation", package = "funfield")`.

The worked example is making a cup of drip coffee: ready your supplies,
turn the machine on, pour, sip. We build that field once and then run it
two ways, which answer two different questions:

- **What does the plan ultimately afford?** Resolve the whole causal
  chain at once and read off the result. This is a *topological sweep*.
- **How does the process unfold in time, step by step, using things
  up as it goes?** Advance the state one move at a time and let earlier
  stages be *consumed* as the token moves down the line. This needs the
  field to carry **memory**, which brings in the **stock / flow**
  distinction.

## The field step

A functional field iterates

\[
F_t = F(s_t), \qquad s_{t+1} = s_t \cdot F_t,
\]

where `s_t` is the situational **state** (a named vector of node values)
and `F_t` is the field evaluated at that state. Two helpers run the
loop:

- [`evalF()`](../reference/evalF.html) takes one step: given the field
  (a `lavaan` model string), its parameter values, and the current
  `s_t`, it returns `s_{t+1}`.
- [`runF()`](../reference/runF.html) drives the iteration for a fixed
  number of steps and returns the full trajectory.

Both take a `mode` argument. `mode = "sweep"` (the default) resolves the
entire causal chain in a single step; `mode = "sync"` advances one edge
per step, which is what we need once the process unfolds in time. We
meet both below.

We build the field in two halves that `lavaan` lets us stitch together
by concatenation: the **situation** — the affordance structure of the
world, which exists whether or not anyone acts — and the **action
plan** — a set of conditional intentions the agent layers on top.

## Step 1: The situation

Each stage of the coffee process is an **afforded force**: a path that
contributes nothing until its prerequisite state variable is on. In the
`fZ_X.Y` schema an afforded force is a term with a moderator but no
baseline — `fZ_X.Y * X:Z` — so the `X → Y` force is zero whenever
`s[Z] = 0` and switches on once `s[Z]` flips positive.

The four gated stages, plus a terminal utility force:

| Force | Afforded by | Reads as |
|---|---|---|
| `make → s2`        | `s1`     | setting up the rig works once you're ready to make coffee |
| `TurnOn → HCoPot`  | `s2`     | turning it on brews into the pot once the rig is ready |
| `Pour → HCoCup`    | `HCoPot` | pouring fills the cup once there's coffee in the pot |
| `Sip → HCo`        | `HCoCup` | sipping yields "having coffee" once it's in the cup |
| `HCo → L`          | —        | having coffee produces likelihood / value (baseline) |

The terminal node is **`L`** — likelihood. In ESJT data `L` is the rated
likelihood of taking an action, and here it plays the role an economist
would give *subjective expected utility*: the field assumes a person
takes the action they appraise most positively. Reading `L` off the
settled field is reading off which action the agent is predicted to
choose.

Node names are lavaan-safe (letters, digits, `_`); the `.` inside each
parameter label is the source/target separator.
[`labelF()`](../reference/labelF.html) derives the `fZ_X.Y` label for
each term and captures any coefficient we fix inline into a matching
`params` vector. The labelling reads condition-first off the right-hand
side — `Y ~ Z:X` becomes `fZ_X.Y` — so each interaction is written
**condition first, action last**. We fix a force of `1` everywhere, so
the wave propagates without attenuation:

```{r situation}
library(funfield)

situation <- labelF("
  s2     ~ 1 * s1:make
  HCoPot ~ 1 * s2:TurnOn
  HCoCup ~ 1 * HCoPot:Pour
  HCo    ~ 1 * HCoCup:Sip
  L      ~ 1 * HCo
", actions = c("make", "TurnOn", "Pour", "Sip"))

cat(situation$model)
situation$params
```

`s2:TurnOn` is the force from the action `TurnOn` to `HCoPot`, afforded
by the state `s2`, labelled `fs2_TurnOn.HCoPot`. The lone baseline term
`L ~ HCo` becomes `f1_HCo.L` — always active once `HCo` is present.
Nothing here actually *fires* an action; the situation says only what
each action *would* produce given the right prerequisites. The firing is
the plan's job.

## Step 2: The action plan

A **conditional action plan** says *when* to act: fire each action once
its prerequisite situational state is achieved. Each rule is a force
from a *state* to an *action* — the opposite direction from the
situation's action → outcome forces:

```{r plan}
plan <- labelF("
  make   ~ 1 * s1
  TurnOn ~ 1 * s2
  Pour   ~ 1 * HCoPot
  Sip    ~ 1 * HCoCup
")

cat(plan$model)
```

Fixing each trigger at `1` is a structural commitment — "act, fully,
once the prerequisite holds" — rather than a force estimated from data.
Because both halves are plain `lavaan` strings, we stitch the field
together and combine the parameter vectors:

```{r stitch}
model  <- paste(situation$model, plan$model, sep = "\n")
params <- c(situation$params, plan$params)
```

## Step 3: Sweep — what the plan affords

The agent starts **ready to make coffee** (`s1 = 1`) and everything else
off. The default `mode = "sweep"` evaluates the nodes in causal order,
each reading the freshly-updated values of its inputs, so the whole
chain resolves in a **single step**:

```{r sweep}
s_0 <- c(s1 = 1, make = 0, s2 = 0, TurnOn = 0, HCoPot = 0,
         Pour = 0, HCoCup = 0, Sip = 0, HCo = 0, L = 0)

round(runF(model, params, s_0, steps = 2), 2)
```

By `t = 1` every node is on and `L = 1`: given the plan and a ready
agent, the field affords coffee. The sweep handles the gating correctly
even though it happens "all at once" — `HCoPot`'s force is keyed on
`s2`, and because `s2` is computed before `HCoPot` in the causal
ordering, the gate is already open by the time it is read. This is the
quick **equilibrium read**: *does this plan, in this situation, afford
the outcome?* — answered in one resolution, with `L` as the verdict.

## Step 4: A process happens in time — and uses things up

The sweep tells us the plan *affords* coffee, but it says nothing about
*sequence* or *resources*. Two things are missing from the affordance
picture if we want to watch the process actually run:

1. **Order in time.** Sipping is not merely afforded after pouring — it
   *happens* after it.
2. **Consumption.** Making coffee uses things up. Once the machine has
   brewed, the grounds and water (`s2`, "ready to make") are *spent*;
   you are no longer "about to make coffee," you have made it.

Neither is captured by forces that are simply afforded-or-not. To run
the process in time we switch to `mode = "sync"` (one move per step) and
give the field two new capacities: **memory**, so a state stays on once
attained, and **consumption**, so an earlier state empties out once the
next stage draws on it. That is the **stock / flow** distinction.

## Step 5: Stocks, flows, and consumption

Read each state's update as a discrete **stock balance**,
`new = old + inflow − outflow`:

- A **stock** *latches*: it carries its previous value forward, and the
  model's terms add to it (inflow, `+`) or subtract from it (outflow /
  consumption, `−`). Object states are stocks — once there is coffee in
  the pot, it stays there until poured.
- A **flow** is *transient*: recomputed from scratch each step, so it is
  on only while its trigger is. Actions are flows — `make` fires while
  you are ready, not forever.
- A **readout** like `L` recomputes too: it reflects current value
  rather than accumulating it.

`lavaan` forbids the self-edge `Y ~ Y`, so the carry-forward that makes a
stock remember cannot be written in the model string — the engine
supplies it when you mark a node a stock. Everything else *is* writable:
production is an ordinary inflow term, and **consumption is an outflow
term with a negative coefficient**, e.g. `use * s2:TurnOn` with
`use = -1` ("turning on uses up the ready state"). The situation gains
those consumption arcs:

```{r situation-tf}
situation_tf <- "
  s1     ~ use * s1:make
  s2     ~ 1 * s1:make     + use * s2:TurnOn
  HCoPot ~ 1 * s2:TurnOn   + use * HCoPot:Pour
  HCoCup ~ 1 * HCoPot:Pour + use * HCoCup:Sip
  HCo    ~ 1 * HCoCup:Sip
  L      ~ 1 * HCo
"
plan_tf <- "
  make   ~ 1 * s1
  TurnOn ~ 1 * s2
  Pour   ~ 1 * HCoPot
  Sip    ~ 1 * HCoCup
"
model_tf  <- paste(situation_tf, plan_tf, sep = "\n")
params_tf <- c(use = -1)
```

We do not hand-list which nodes latch. The split is structural — the
situation's left-hand sides are the **object states** (stocks) and the
plan's are the **actions** (flows) — so passing `actions =` lets
`runF()` derive the stock set automatically: actions and the `L` readout
recompute, everything else latches.

```{r run-tf}
traj_tf <- runF(model_tf, params_tf, s_0,
                actions = c("make", "TurnOn", "Pour", "Sip"),
                mode = "sync", steps = 12)
round(traj_tf, 2)
```

This is a single **token** flowing down the line. Each object state is
produced, persists for a step, then is consumed by the next stage and
returns to `0`; each action fires as a brief pulse and dies once the
state that triggered it is spent. Only the terminal `HCo` and its
readout `L` stay on — you keep the coffee. Compare with the sweep, which
collapsed all of this into the single question "is the outcome
afforded?"; here we see the affordance being *realised*, one consumed
resource at a time.

### The same wave as a heatmap

The diagonal structure reads cleanly as a stage × time tile plot:

```{r wave, eval=has_ggplot, fig.alt="Tile plot of state value by process stage and time step, showing a diagonal wave that lights each stage briefly before it is consumed."}
library(ggplot2)

stage_order  <- c("s1","make","s2","TurnOn","HCoPot",
                  "Pour","HCoCup","Sip","HCo","L")
stage_labels <- c(s1="ready", make="make", s2="rig set", TurnOn="turn on",
                  HCoPot="coffee in pot", Pour="pour", HCoCup="coffee in cup",
                  Sip="sip", HCo="have coffee", L="L")

long <- data.frame(
  step  = rep(0:12, times = length(stage_order)),
  stage = factor(rep(stage_order, each = nrow(traj_tf)),
                 levels = rev(stage_order)),
  value = as.vector(traj_tf[, stage_order])
)

ggplot(long, aes(step, stage, fill = value)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  scale_fill_gradient(low = "white", high = "#1572da", limits = c(0, 1)) +
  scale_x_continuous(breaks = 0:12, expand = c(0, 0)) +
  scale_y_discrete(labels = stage_labels) +
  labs(x = "time step", y = NULL, fill = "state",
       title = "A token flowing down the coffee field") +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank(), legend.position = "right")
```

The lit cells march up and to the right, each stage glowing for a step
or two before its successor consumes it — the visual signature of a
process that uses its inputs up as it proceeds.

## Step 6: Watch the token climb

[`plotField()`](../reference/plotField.html) draws the field at one
state as a node-link diagram: **object** stocks are diamonds, **action**
flows are right-facing triangles, and the terminal **likelihood** node
is a left-facing triangle. Each node is shaded by its current value
(white at `0`, blue at `1`), and an edge is drawn for every currently
afforded force. Passing `plan = plan_tf` colours the plan's forces
**gold**, distinct from the situation's **grey**.

```{r layout}
layout <- data.frame(
  name  = c("s1","make","s2","TurnOn","HCoPot",
            "Pour","HCoCup","Sip","HCo","L"),
  x     = c(0,    2,     3,   5,       6,
            8,     9,      11,   12,   14),
  y     = c(0,    0,     2,   2,       4,
            4,     6,      6,    8,    8),
  shape = c("diamond","rtTri","diamond","rtTri","diamond",
            "rtTri","diamond","rtTri","diamond","lfTri"),
  label = c("ready","make","rig\nset","turn on","coffee\nin pot",
            "pour","coffee\nin cup","sip","have\ncoffee","L"),
  stringsAsFactors = FALSE
)
```

Stepping `plotField()` across the trajectory and stitching the frames
with [`plotsAsWidget()`](../reference/plotsAsWidget.html) animates the
token. Use **← Back** / **Forward →** to watch each diamond fill as its
stage is reached and empty again as the next stage consumes it, the lit
node climbing the staircase:

```{r field-widget, eval=has_ggplot, fig.alt="Interactive staircase field diagram stepping through the coffee simulation; a single lit node climbs the staircase as each stage is reached and then consumed."}
frames <- lapply(0:12, function(t) {
  plotField(model_tf, params_tf, traj_tf[paste0("t=", t), ], layout,
            plan = plan_tf, title = paste("t =", t))
})

plotsAsWidget(
  frames,
  panel_titles = paste("Step", 0:12),
  width = 9, height = 6
)
```

## Out-of-bounds means a misspecified field

`runF()` does **not** clamp. Functional-field state is expected to stay
within `[-1, 1]` when the model is well specified, so a value outside
that range is treated as a signal that something is wrong rather than
quietly truncated. The clearest way to trip it is to give a stock an
inflow but no consumption — it then accumulates without bound:

```{r runaway, warning=TRUE}
runaway <- runF("HCoPot ~ 1 * s2:TurnOn", params = c(use = 0),
                s_0 = c(s2 = 1, TurnOn = 1, HCoPot = 0),
                stocks = "HCoPot", mode = "sync", steps = 4)
runaway[, "HCoPot"]
```

The warning names the offending node and step, and the true (un-clamped)
value is returned so the overshoot is visible. In the full coffee model
the consumption terms are exactly what keep every stock bounded: each
stage hands its token on and empties, so nothing piles up.

## Takeaway

One field, run two ways. The **sweep** answers the affordance question
in a single resolution — *given this situation and plan, is the outcome
reachable?* — with `L` as the verdict. The **sync** run with stocks and
flows answers the process question — *how does it unfold, and what gets
used up along the way?* — showing a token move down the chain as each
resource is produced, held, and consumed.

The division of labour stays explicit throughout: the **plan** (gold)
holds standing intentions keyed on states, the **situation** (grey)
decides whether a fired action produces anything, and the **stock /
flow** typing decides what persists. You can intend to turn the machine
on from the very first step, yet nothing brews until the rig is set up —
and once it has brewed, the ready state is spent and that intention has
nothing left to act on.

This is the deductive use of a functional field. The inductive
direction — recovering the situation's coefficients from data — is the
subject of the `pathXMY` vignettes
(`vignette("speeding_pathXMY", package = "funfield")` and its
companions).
```
