---
title: "Coffee Field Model: Sequential Action Planning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Coffee Field Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 9,
  fig.height = 5
)
has_qgraph <- requireNamespace("qgraph", quietly = TRUE)
```

## What Is a Functional Field?

A **functional field** is a model of how situational forces propagate through
an environment to produce action outcomes.  At each moment the field specifies
which forces are *afforded* — i.e., which causal pathways are open given the
current state of the situation.  Forces spread step-by-step through the network:
disturbances at time *t* are transmitted through the field matrix to produce
new disturbances at time *t + 1*, which then accumulate into the situation.

This vignette walks through a concrete example — making a cup of drip coffee —
to show how a four-step action sequence becomes causally connected through a
field model.  The key insight is **causal gating**: later steps are only
*afforded* once earlier steps have changed the situation.  A plan adds
intentional impulses that drive the sequence forward, but the outcome of each
step is what opens up the next pathway.

---

## Step 1: Define the Layout

The layout is a character matrix where each non-zero cell encodes one node as
`"type_name"`.  See `?prepField` for the full list of type codes; the ones used
here are:

| Code | Meaning |
|------|---------|
| `a`  | Appraisal (left-facing triangle) |
| `o`  | Object — binary present / absent (diamond) |
| `d`  | Do / decision action (right-facing triangle) |

The "4 step DC to mouth" layout uses a **staircase** arrangement: each
action–outcome pair steps one row up and two columns to the right.  This
diagonal flow naturally represents the sequential, irreversible character of
the process — you cannot sip until you have poured, pour until it has brewed,
etc.  The motivational prerequisites (`Want(Co).p`, `H(Ings).p`) and the
terminal utility appraisal (`U.p`) are placed in the rightmost column.

```{r layout}
# During development, load from source so new functions are always available.
# Replace with library(funfield) once the package is installed.
pkgload::load_all("..", quiet = TRUE)

coffee_layout <- matrix(0, nrow = 5, ncol = 15)

# --- Staircase: bottom-left to top-right ---
coffee_layout[5,  1] <- "o_s1"              # combined readiness state
coffee_layout[5,  3] <- "d_make(s2).p"      # action: set up the drip coffeemaker
coffee_layout[4,  4] <- "o_s2"              # coffeemaker is set up and ready
coffee_layout[4,  6] <- "d_turn(ON.DC).p"   # action: turn on the drip coffeemaker
coffee_layout[3,  7] <- "o_H(Co).Pot"       # coffee has brewed into the pot
coffee_layout[3,  9] <- "d_pour(Pot,Cup).p" # action: pour from pot into cup
coffee_layout[2, 10] <- "o_H(Co).Cup"       # coffee is in the cup
coffee_layout[2, 12] <- "d_sip(Cup).p"      # action: sip from the cup
coffee_layout[1, 13] <- "o_H(Co).p"         # having / enjoyed the coffee

# --- Right column: motivational inputs and terminal utility ---
coffee_layout[1, 15] <- "a_U.p"             # appraisal: utility / satisfaction
coffee_layout[4, 15] <- "o_Want(Co).p"      # wanting coffee
coffee_layout[5, 15] <- "o_H(Ings).p"       # having the necessary ingredients

fprops <- prepField(coffee_layout)

# Full English display labels (internal names → readable text)
coffee_labels <- c(
  "s1"              = "Ready to\nmake coffee",
  "make(s2).p"      = "Set up\ncoffeemaker",
  "s2"              = "Coffeemaker\nready",
  "turn(ON.DC).p"   = "Turn on\ncoffeemaker",
  "H(Co).Pot"       = "Coffee\nin pot",
  "pour(Pot,Cup).p" = "Pour into\ncup",
  "H(Co).Cup"       = "Coffee\nin cup",
  "sip(Cup).p"      = "Sip from\ncup",
  "H(Co).p"         = "Have\ncoffee",
  "U.p"             = "Satisfied",
  "Want(Co).p"      = "Want\ncoffee",
  "H(Ings).p"       = "Have\ningredients"
)
```

`prepField()` strips the type prefix and returns `fprops`, a list that bundles
the processed layout, node names, shapes, sizes, and an empty field matrix
(`fprops$fmat`) ready to be filled with paths.

You can optionally inspect or adjust the layout interactively:

```{r editor, eval=FALSE}
# Opens a drag-and-drop editor — not run during vignette build
fieldEditor(coffee_layout)
```

---

## Step 2: Define the Field Function

`ff()` computes the force-transmission matrix `ft` from the *current*
situation vector `s`.  Two kinds of paths populate `ft`:

1. **Afforded paths** — only present when a prerequisite situation node is
   active.  For example, the path `make(s2).p → s2` is only afforded while
   `s["s1"] > 0`.  If the prerequisite is absent, the path weight is zero and
   no force flows — the gate is closed.

2. **Planned paths** — the agent's goal-directed intention to act at each
   state.  These are supplied by the `plan` argument and added on top of
   afforded paths.

```{r ff}
ff <- function(s, plan = DoNothing) {
  ft <- fprops$fmat   # start with the zero matrix (names already set)

  # --- Afforded paths (gated by situation state) ---

  # Gate 1: s1 present → setting up the coffeemaker produces s2
  if (s["s1"] > 0) {
    ft["make(s2).p", "s2"] <- 0.9
  }

  # Gate 2: coffeemaker ready (s2) → turning it on brews coffee into the pot
  if (s["s2"] > 0) {
    ft["turn(ON.DC).p", "H(Co).Pot"] <- 0.9
  }

  # Gate 3: coffee in the pot → pouring fills the cup
  if (s["H(Co).Pot"] > 0) {
    ft["pour(Pot,Cup).p", "H(Co).Cup"] <- 0.9
  }

  # Gate 4: coffee in the cup → sipping produces the "have coffee" state
  if (s["H(Co).Cup"] > 0) {
    ft["sip(Cup).p", "H(Co).p"] <- 0.9
  }

  # Terminal: having coffee produces utility
  if (s["H(Co).p"] > 0) {
    ft["H(Co).p", "U.p"] <- 0.9
  }

  # --- Plan forces (intentional action impulses) ---
  # plan() returns a matrix with 1s where the agent intends to act.
  # Adding these forces lets disturbances from situation nodes drive
  # the corresponding action nodes forward.
  ft <- ft + plan(ft)

  list(ft = ft)
}
```

### Why causal gating works here

At time step 1 the situation is `[s1=1, Want(Co).p=1, H(Ings).p=1, ...]`.
Because `s["s2"] = 0`, the gate for `turn(ON.DC).p → H(Co).Pot` is closed:
that path has weight zero and no force flows through it.  Even though the agent
plans to turn on the coffeemaker, the outcome of doing so is not yet causally
possible — the coffeemaker has not been set up yet.  Only once `s2` becomes
positive (the coffeemaker is ready) does that gate open, and force can
propagate to the next stage.

---

## Step 3: Define Plans

A **plan** is a function that accepts a field matrix and returns a same-shaped
matrix marking which paths the agent *intends* to use (coded as 1).
`DoNothing` returns all zeros; `do4step` adds one planned impulse per step,
connecting each situation outcome to its corresponding next action.

```{r plans}
DoNothing <- function(x) { x * 0 }

do4step <- function(x) {
  x[] <- 0
  x["s1",        "make(s2).p"]      <- 1  # in s1: set up the coffeemaker
  x["s2",        "turn(ON.DC).p"]   <- 1  # ready: turn it on
  x["H(Co).Pot", "pour(Pot,Cup).p"] <- 1  # brewed: pour into cup
  x["H(Co).Cup", "sip(Cup).p"]      <- 1  # in cup: sip
  x
}
```

---

## Step 4: Set Up the Initial Situation and Run the Simulation

The initial situation `s` is all zeros.  The *action* (second argument to
`expOutcomesFF`) sets the starting disturbances: the agent wants coffee
(`Want(Co).p`), has the ingredients (`H(Ings).p`), and the combined readiness
state `s1` is present.

```{r simulate}
# Initial situation: everything at zero
s <- setNames(rep(0, length(fprops$vnames)), as.vector(fprops$vnames))

# Run the simulation with the four-step plan
sitlog <- expOutcomesFF(
  s,
  i      = c("s1", "Want(Co).p", "H(Ings).p"),
  ffield = ff,
  plan   = do4step
)

# Quick look at the situation matrix (rows = time steps, cols = nodes)
round(sitlog$sit[1:9, ], 2)
```

---

## Step 5: Visualize with `fieldSequenceWidget()`

`fieldSequenceWidget()` renders each time step as a high-resolution image and
wraps them in an interactive widget with **&#8592; Back** and **Forward &#8594;**
buttons, so you can step through the simulation one frame at a time.  Images
are rendered at 2× the display size for sharpness on retina screens.

Here we select the four key steps — the moments just *after* each new gate
opens:

```{r widget, eval=has_qgraph}
fieldSequenceWidget(
  sitlog, fprops,
  steps       = c(1, 3, 5, 7),
  node_labels = coffee_labels,
  titles = c(
    "Step 1: Gate 1 open \u2014 set up coffeemaker",
    "Step 3: Gate 2 open \u2014 turn on; coffee brews into pot",
    "Step 5: Gate 3 open \u2014 pour pot into cup",
    "Step 7: Gate 4 open \u2014 sip; coffee acquired"
  )
)
```

To step through every simulated time step, omit `steps` entirely:

```{r widget_all, eval=has_qgraph}
fieldSequenceWidget(sitlog, fprops, node_labels = coffee_labels)
```

---

## Interpretation

Each plot shows the *active* field at that moment.  Edges appear only where
the affordance gate is open (the prerequisite situation node is positive) and,
for planned actions, an intentional impulse is also present:

| Step | Gate opens when...    | Path that becomes active |
|------|-----------------------|--------------------------|
| 1    | `s1` present          | `s1 → make(s2).p → s2` |
| 3    | `s2` present          | `s2 → turn(ON.DC).p → H(Co).Pot` |
| 5    | `H(Co).Pot` present   | `H(Co).Pot → pour(Pot,Cup).p → H(Co).Cup` |
| 7    | `H(Co).Cup` present   | `H(Co).Cup → sip(Cup).p → H(Co).p` |
| 9+   | `H(Co).p` present     | `H(Co).p → U.p` (terminal utility) |

**Key takeaway:** The staircase layout makes the sequential dependency
tangible.  Sipping is not merely *last* in time — it is only *causally
possible* once coffee is in the cup, which requires having poured from the
pot, which requires having brewed, which requires setting up the machine.
The plan supplies the motivational impulses at each stage, but the affordance
structure of the situation controls which actions can have downstream effects
at any given moment.
