---
title: "Coffee Field Model: Affordances, Choice, and a Process in Time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Coffee Field Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4
)
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
```

## What is a functional field?

A **functional field** is a model of how situational forces propagate
through an environment to produce action outcomes. At each moment the
field specifies which forces are *afforded* — which causal pathways are
open given the current state of the situation. A later step only matters
once an earlier step has changed the situation: this is **causal
gating**, and it is what makes a four-step plan into a connected causal
chain rather than four independent intentions.

The same field `F` supports two complementary workflows. **Induction**
runs data → `F`: you fit the field's coefficients from observed
behaviour with `pathXMY()` and friends. **Deduction** runs `F` →
behaviour: you treat the coefficients as a working theory and *simulate*
forward to see what a plan produces. This vignette is about deduction.
For the notation it uses throughout — the `F` / `F_t` field objects and
the `fZ_X.Y` parameter schema — see
`vignette("notation", package = "funfield")`.

The worked example is making a cup of drip coffee: ready your supplies,
turn the machine on, pour, sip. We build that field once and put it to
three uses, which answer three different questions:

- **What does a plan afford?** Resolve its whole causal chain at once and
  read off the result. This is a *topological sweep*.
- **Which plan should the agent choose?** Appraise several candidate
  plans the same way and keep the one that affords the most value. This
  is what a **choice / chance node** does, and
  [`chooseF()`](../reference/chooseF.html) computes it.
- **How does the chosen plan unfold in time, using things up as it
  goes?** Advance the state one move at a time and let earlier stages be
  *consumed* as the token moves down the line. This needs the field to
  carry **memory**, which brings in the **stock / flow** distinction.

## The field step

A functional field iterates

\[
F_t = F(s_t), \qquad s_{t+1} = s_t \cdot F_t,
\]

where `s_t` is the situational **state** (a named vector of node values)
and `F_t` is the field evaluated at that state. Two helpers run the
loop:

- [`evalF()`](../reference/evalF.html) takes one step: given the field
  (a `lavaan` model string), its parameter values, and the current
  `s_t`, it returns `s_{t+1}`.
- [`runF()`](../reference/runF.html) drives the iteration for a fixed
  number of steps and returns the full trajectory.

Both take a `mode` argument. `mode = "sweep"` (the default) resolves the
entire causal chain in a single step; `mode = "sync"` advances one edge
per step, which is what we need once the process unfolds in time. We
meet both below.

We build the field in two halves that `lavaan` lets us stitch together
by concatenation: the **situation** — the affordance structure of the
world, which exists whether or not anyone acts — and the **action
plan** — a set of conditional intentions the agent layers on top.

## Step 1: The situation and the choice point

Each stage of the coffee process is an **afforded force**: a path that
contributes nothing until its prerequisite state variable is on. In the
`fZ_X.Y` schema an afforded force is a term with a moderator but no
baseline — `fZ_X.Y * X:Z` — so the `X → Y` force is zero whenever
`s[Z] = 0` and switches on once `s[Z]` flips positive.

The physical chain, plus a terminal value force:

| Force | Afforded by | Reads as |
|---|---|---|
| `make → s2`        | —        | making the rig sets it up (the entry point) |
| `TurnOn → HCoPot`  | `s2`     | turning it on brews into the pot once the rig is ready |
| `Pour → HCoCup`    | `HCoPot` | pouring fills the cup once there's coffee in the pot |
| `Sip → HCo`        | `HCoCup` | sipping yields "having coffee" once it's in the cup |
| `HCo → L`          | —        | having (consumed) coffee produces likelihood / value |

The terminal node is **`L`** — likelihood. In ESJT data `L` is the rated
likelihood of taking an action, and here it plays the role an economist
would give *subjective expected utility*: the field assumes a person
takes the action they appraise most positively. Reading `L` off the
settled field is reading off how good an action looks.

That lone baseline `HCo → L` carries the model's whole motivational
charge. A strong `f1_HCo.L` says *if an action resolves with me having
consumed coffee, I am much more likely to do it* — which is just what it
means, here, to **want coffee**. Wanting is not a node; it is a property
of the situation, the standing force that gives the terminal outcome its
value. Everything upstream is causally neutral plumbing that matters
only because of where it leads.

Node names are lavaan-safe (letters, digits, `_`); the `.` inside each
parameter label is the source/target separator.
[`labelF()`](../reference/labelF.html) derives the `fZ_X.Y` label for
each term and captures any coefficient we fix inline into a matching
`params` vector. The labelling reads condition-first off the right-hand
side — `Y ~ Z:X` becomes `fZ_X.Y` — so each interaction is written
**condition first, action last**. We fix a force of `1` everywhere, so
the wave propagates without attenuation:

```{r situation}
library(funfield)

situation <- labelF("
  s2     ~ 1 * make
  HCoPot ~ 1 * s2:TurnOn
  HCoCup ~ 1 * HCoPot:Pour
  HCo    ~ 1 * HCoCup:Sip
  L      ~ 1 * HCo
", actions = c("make", "TurnOn", "Pour", "Sip"))

cat(situation$model)
situation$params
```

`s2:TurnOn` is the force from the action `TurnOn` to `HCoPot`, afforded
by the state `s2`, labelled `fs2_TurnOn.HCoPot`. The lone baseline term
`L ~ HCo` becomes `f1_HCo.L` — the want — and `make → s2` is a baseline
too (`f1_make.s2`): making the rig is the entry point, gated by nothing
but the act itself. Nothing here actually *fires* an action; the
situation says only what each action *would* produce given the right
prerequisites.

Sitting at the head of the field is the **choice point** (a
*choice / chance node*, which we draw as a circle). Its job is not to
want anything but to **appraise the whole situation and pick a
response**: it weighs the candidate plans by simulating what each
affords and commits to the one with the highest expected value, with `L`
standing in for subjective expected utility. The wanting lives in the
`f1_HCo.L` force; the choosing lives here. We meet the mechanism in
Step 3.

## Step 2: Candidate action plans

A **conditional action plan** says *when* to act: fire each action once
its prerequisite situational state is achieved. Each rule is a force
from a *state* to an *action* — the opposite direction from the
situation's action → outcome forces. The first action is triggered not
by a situational state but by the **choice point** itself: committing to
the plan is what sets it in motion.

```{r plan-full}
full <- "
  make   ~ 1 * choice
  TurnOn ~ 1 * s2
  Pour   ~ 1 * HCoPot
  Sip    ~ 1 * HCoCup
"
```

To give the choice point something to choose between, we draft a second
plan that skips a step — turning the machine on and then reaching for a
sip without pouring first:

```{r plan-nopour}
no_pour <- "
  make   ~ 1 * choice
  TurnOn ~ 1 * s2
  Sip    ~ 1 * HCoCup
"
```

Fixing each trigger at `1` is a structural commitment — "act, fully,
once the prerequisite holds" — rather than a force estimated from data.

## Step 3: Appraise and choose

The agent starts at the choice point, committed (`choice = 1`), with
everything else off. Stitch a plan onto the situation and the default
`mode = "sweep"` evaluates the nodes in causal order, each reading the
freshly-updated values of its inputs, so the whole chain resolves in a
**single step**:

```{r sweep-one}
s_0 <- c(choice = 1, make = 0, s2 = 0, TurnOn = 0, HCoPot = 0,
         Pour = 0, HCoCup = 0, Sip = 0, HCo = 0, L = 0)

model  <- paste(situation$model, full, sep = "\n")
round(runF(model, situation$params, s_0, steps = 1), 2)
```

By `t = 1` the whole chain has settled and `L = 1`: given the full plan
and a committed agent, the field affords coffee. The sweep handles the
gating correctly even though it happens "all at once" — `HCoPot`'s force
is keyed on `s2`, and because `s2` is computed before `HCoPot` in the
causal ordering, the gate is already open by the time it is read. This
is the **equilibrium read**: *does this plan, in this situation, reach
the outcome?* — answered in one resolution, with `L` as the verdict.

A choice point needs at least two responses to choose between.
[`chooseF()`](../reference/chooseF.html) appraises each candidate plan
with exactly this sweep, tabulates what it affords, and flags the
highest-value one. It appends a do-nothing baseline by default, giving
the minimal "act vs. do nothing" contrast:

```{r choose}
chooseF(situation$model, situation$params, s_0,
        plans = list(full = full, no_pour = no_pour))
```

The full plan reaches `HCo = 1` and earns `L = 1`; doing nothing earns
`L = 0`. The telling row is `no_pour`. It turns the machine on —
`HCoPot = 1`, there really is coffee in the pot — but because the
`Sip → HCo` force is gated on `HCoCup`, and the cup is never filled,
sipping affords nothing: `HCoCup`, `HCo`, and `L` all stay at `0`. A
plan that skips a prerequisite step reaches a dead end, and the choice
point reads that straight off the value column and selects the full
plan. Causal gating is doing the work: a plan is only as good as its
weakest link in the chain.

## Step 4: A process happens in time — and uses things up

The sweep tells us the chosen plan *affords* coffee, but it says nothing
about *sequence* or *resources*. Two things are missing from the
affordance picture if we want to watch the process actually run:

1. **Order in time.** Sipping is not merely afforded after pouring — it
   *happens* after it.
2. **Consumption.** Making coffee uses things up. Once the machine has
   brewed, the rig (`s2`, the set-up state) is *spent*; you are no
   longer "about to brew," you have brewed.

Neither is captured by forces that are simply afforded-or-not. To run
the process in time we switch to `mode = "sync"` (one move per step) and
give the field two new capacities: **memory**, so a state stays on once
attained, and **consumption**, so an earlier state empties out once the
next stage draws on it. That is the **stock / flow** distinction.

## Step 5: Stocks, flows, and a one-shot choice

Read each state's update as a discrete **stock balance**,
`new = old + inflow − outflow`:

- A **stock** *latches*: it carries its previous value forward, and the
  model's terms add to it (inflow, `+`) or subtract from it (outflow /
  consumption, `−`). Object states are stocks — once there is coffee in
  the pot, it stays there until poured.
- A **flow** is *transient*: recomputed from scratch each step, so it is
  on only while its trigger is. Actions are flows — `make` fires while
  the choice fires it, not forever.
- A **readout** like `L` recomputes too: it reflects current value
  rather than accumulating it.

`lavaan` forbids the self-edge `Y ~ Y`, so the carry-forward that makes a
stock remember cannot be written in the model string — the engine
supplies it when you mark a node a stock. Everything else *is* writable:
production is an ordinary inflow term, and **consumption is an outflow
term with a negative coefficient**, e.g. `use * s2:TurnOn` with
`use = -1` ("turning on uses up the set-up rig"). The situation gains
those consumption arcs:

```{r situation-tf}
situation_tf <- "
  s2     ~ 1 * make        + use * s2:TurnOn
  HCoPot ~ 1 * s2:TurnOn   + use * HCoPot:Pour
  HCoCup ~ 1 * HCoPot:Pour + use * HCoCup:Sip
  HCo    ~ 1 * HCoCup:Sip
  L      ~ 1 * HCo
"
model_tf  <- paste(situation_tf, full, sep = "\n")
params_tf <- c(use = -1)
```

The **choice point** is a flow too — and a special one. A decision is an
event, not a level: once the agent commits and fires the first action,
the choice is **spent**. It does not persist the way "being ready"
would, and it is not consumed later by a downstream stage; it simply
**depletes on activation**. Because `choice` never appears on a
left-hand side — nothing in the field *produces* a decision — we mark it
a flow so the engine treats it as a **one-shot pulse**: seeded at the
start, read once by the action it triggers, then emptied.

```{r run-tf}
traj_tf <- runF(model_tf, params_tf, s_0,
                actions = c("make", "TurnOn", "Pour", "Sip"),
                flows   = "choice",
                mode = "sync", steps = 12)
round(traj_tf, 2)
```

We do not hand-list which object states latch. The split is structural —
the situation's left-hand sides are the **object states** (stocks) and
the plan's are the **actions** (flows) — so passing `actions =` lets
`runF()` derive the stock set automatically: actions and the `L` readout
recompute, everything else latches. `flows = "choice"` adds the one
exogenous node that should fire once and deplete rather than persist.

This is a single **token** flowing down the line, kicked off by the
choice point. The circle lights once at `t = 0` and is dark by `t = 1` —
the decision, made and spent. Each object state is then produced,
persists for a step, then is consumed by the next stage and returns to
`0`; each action fires as a brief pulse and dies once the state that
triggered it is spent. Only the terminal `HCo` and its readout `L` stay
on — you keep the coffee. Compare with the sweep, which collapsed all of
this into the single question "is the outcome afforded?"; here we see the
affordance being *realised*, one consumed resource at a time.

### The same wave as a heatmap

The diagonal structure reads cleanly as a stage × time tile plot:

```{r wave, eval=has_ggplot, fig.alt="Tile plot of state value by process stage and time step, showing a diagonal wave that lights each stage briefly before it is consumed."}
library(ggplot2)

stage_order  <- c("choice","make","s2","TurnOn","HCoPot",
                  "Pour","HCoCup","Sip","HCo","L")
stage_labels <- c(choice="choice", make="make", s2="rig set", TurnOn="turn on",
                  HCoPot="coffee in pot", Pour="pour", HCoCup="coffee in cup",
                  Sip="sip", HCo="have coffee", L="L")

long <- data.frame(
  step  = rep(0:12, times = length(stage_order)),
  stage = factor(rep(stage_order, each = nrow(traj_tf)),
                 levels = rev(stage_order)),
  value = as.vector(traj_tf[, stage_order])
)

ggplot(long, aes(step, stage, fill = value)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  scale_fill_gradient(low = "white", high = "#1572da", limits = c(0, 1)) +
  scale_x_continuous(breaks = 0:12, expand = c(0, 0)) +
  scale_y_discrete(labels = stage_labels) +
  labs(x = "time step", y = NULL, fill = "state",
       title = "A token flowing down the coffee field") +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank(), legend.position = "right")
```

The lit cells march up and to the right: the choice point flashes once at
the bottom-left, then each stage glows for a step or two before its
successor consumes it — the visual signature of a process that uses its
inputs up as it proceeds.

## Step 6: Watch the token climb

[`plotField()`](../reference/plotField.html) draws the field at one
state as a node-link diagram: the **choice point** is a circle,
**object** stocks are diamonds, **action** flows are right-facing
triangles, and the terminal **likelihood** node is a left-facing
triangle. Each node is shaded by its current value (white at `0`, blue at
`1`), and an edge is drawn for every currently afforded force. Passing
`plan = full` colours the plan's forces **gold**, distinct from the
situation's **grey**.

```{r layout}
layout <- data.frame(
  name  = c("choice","make","s2","TurnOn","HCoPot",
            "Pour","HCoCup","Sip","HCo","L"),
  x     = c(0,    2,     3,   5,       6,
            8,     9,      11,   12,   14),
  y     = c(0,    0,     2,   2,       4,
            4,     6,      6,    8,    8),
  shape = c("circle","rtTri","diamond","rtTri","diamond",
            "rtTri","diamond","rtTri","diamond","lfTri"),
  label = c("choice","make","rig\nset","turn on","coffee\nin pot",
            "pour","coffee\nin cup","sip","have\ncoffee","L"),
  stringsAsFactors = FALSE
)
```

Stepping `plotField()` across the trajectory and stitching the frames
with [`plotsAsWidget()`](../reference/plotsAsWidget.html) animates the
token. Use **← Back** / **Forward →** to watch the choice point flash and
empty, then each diamond fill as its stage is reached and empty again as
the next stage consumes it, the lit node climbing the staircase:

```{r field-widget, eval=has_ggplot, fig.alt="Interactive staircase field diagram stepping through the coffee simulation; the choice point fires once at the foot of the staircase, then a single lit node climbs as each stage is reached and then consumed."}
frames <- lapply(0:12, function(t) {
  plotField(model_tf, params_tf, traj_tf[paste0("t=", t), ], layout,
            plan = full, title = paste("t =", t))
})

plotsAsWidget(
  frames,
  panel_titles = paste("Step", 0:12),
  width = 9, height = 6
)
```

## Out-of-bounds means a misspecified field

`runF()` does **not** clamp. Functional-field state is expected to stay
within `[-1, 1]` when the model is well specified, so a value outside
that range is treated as a signal that something is wrong rather than
quietly truncated. The clearest way to trip it is to give a stock an
inflow but no consumption — it then accumulates without bound:

```{r runaway, warning=TRUE}
runaway <- runF("HCoPot ~ 1 * s2:TurnOn", params = c(use = 0),
                s_0 = c(s2 = 1, TurnOn = 1, HCoPot = 0),
                stocks = "HCoPot", mode = "sync", steps = 4)
runaway[, "HCoPot"]
```

The warning names the offending node and step, and the true (un-clamped)
value is returned so the overshoot is visible. In the full coffee model
the consumption terms are exactly what keep every stock bounded: each
stage hands its token on and empties, so nothing piles up.

## Takeaway

One field, three uses. The **sweep** answers the affordance question for
a single plan in one resolution — *given this situation and plan, is the
outcome reachable?* — with `L` as the verdict.
[`chooseF()`](../reference/chooseF.html) turns that into a **choice**:
appraise every candidate plan the same way and keep the one whose `L` is
highest, which is all a **choice / chance node** is — an argmax over what
each response affords. And the **sync** run with stocks and flows shows
the chosen plan *unfold* — a token moving down the chain as each resource
is produced, held, and consumed.

The division of labour stays explicit throughout: the **want** is a
property of the situation (the strong `f1_HCo.L` force), the **choice
point** (circle) weighs responses against it, the **plan** (gold) holds
standing intentions keyed on states, the **situation** (grey) decides
whether a fired action produces anything, and the **stock / flow** typing
decides what persists. A plan that skips pouring affords no coffee, so it
is never chosen; and once the chosen plan runs, the choice that launched
it is spent on the very first move.

This is the deductive use of a functional field. The inductive
direction — recovering the situation's coefficients from data — is the
subject of the `pathXMY` vignettes
(`vignette("speeding_pathXMY", package = "funfield")` and its
companions).
```
