---
title: "Coffee Field Model: One Field, One Policy, One Run"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Coffee Field Model: One Field, One Policy, One Run}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4
)
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
```

## What is a functional field?

A **functional field** is a model of how situational forces propagate
through an environment to produce action outcomes. At each moment the
field specifies which forces are *afforded* — which causal pathways are
open given the current state of the situation. A later step only matters
once an earlier step has changed the situation: this is **causal
gating**, and it is what turns a four-step plan into a connected causal
chain rather than four independent intentions.

The same field `F` supports two complementary workflows. **Induction**
runs data → `F`: you fit the field's coefficients from observed
behaviour with `pathXMY()` and friends. **Deduction** runs `F` →
behaviour: you treat the coefficients as a working theory and *simulate*
forward to see what a plan produces. This vignette is about deduction.
For the notation it uses throughout — the `F` field object and the
`fZ_X.Y` parameter schema — see
`vignette("notation", package = "funfield")`.

The worked example is making a cup of drip coffee: ready your supplies,
turn the machine on, pour, sip. The whole vignette rests on one clean
separation:

- the **field** is the *physics* — how actions and states move other
  states and outcomes, including what each action *uses up* and what it
  *costs*. It exists whether or not anyone acts.
- the **policy** is the *plan* — which action the agent is willing to
  take, and when.

[`simulateF()`](../reference/simulateF.html) runs a policy on a field as
a **turn-based process**: each turn it finds the actions whose
conditions are met, fires **one** of them, lets its effects settle, and
moves on — until no action is eligible. Because actions enter one at a
time, the *same run* answers both questions you can ask of a plan: **what
does it ultimately afford?** (read the final value) and **how does it
unfold in time?** (read the whole trajectory).

## Step 1: The field

Each stage of the coffee process is an **afforded force**: a path that
contributes nothing until its prerequisite state is on. In the `fZ_X.Y`
schema that is a term with a moderator but no baseline — `fZ_X.Y * X:Z` —
so the `X → Y` force is zero whenever `s[Z] = 0` and switches on once
`s[Z]` turns positive.

The physical chain of afforded forces:

| Force | Afforded by | Reads as |
|---|---|---|
| `make → s2`        | —        | making the rig sets it up (the entry point) |
| `TurnOn → HCoPot`  | `s2`     | turning it on brews into the pot once the rig is ready |
| `Pour → HCoCup`    | `HCoPot` | pouring fills the cup once there's coffee in the pot |
| `Sip → HCo`        | `HCoCup` | sipping yields "having coffee" once it's in the cup |

A real process also **uses things up**: once the machine has brewed, the
set-up rig (`s2`) is spent; once you pour, the coffee leaves the pot. We
write each consumption as an **outflow term with a shared negative
coefficient** `use`, e.g. `use * s2:TurnOn` ("turning on consumes the
rig"). Production and consumption together make every object state a
balance: it rises when its action produces it and falls when the next
action draws it down.

Two more forces close the field: **value** and **cost**.

The terminal node is **`L`** — likelihood. In ESJT data `L` is the rated
likelihood of taking an action; here it plays the role an economist would
give *subjective expected utility*. Value enters through the forces into
`L`, and we give it two:

| Force | Coefficient | Reads as |
|---|---|---|
| `HCo → L`    | `.9` | having consumed coffee is most of what the agent is after |
| `Energy → L` | `.1` | all else equal, the agent would rather keep its energy |

The vector `f[., L]` — every force into `L` — is the agent's **values**,
its most direct utility function. We keep it normalised so the absolute
weights sum to one (`.9 + .1 = 1`), reading as a budget over what the
agent cares about.

Cost enters through a new node, **`Energy`**. Every action carries a
**performance cost**: doing something — even something futile — spends
effort. [`costF()`](../reference/costF.html) writes that in for us, a
flat decrement to `Energy` per action, charged on the bare action node so
it is paid whenever the action fires, *whether or not the action
accomplishes anything*. Pouring an empty pot still costs effort.

```{r field}
library(funfield)

actions <- c("make", "TurnOn", "Pour", "Sip")

afford <- "
  s2     ~ 1 * make        + use * s2:TurnOn
  HCoPot ~ 1 * s2:TurnOn   + use * HCoPot:Pour
  HCoCup ~ 1 * HCoPot:Pour + use * HCoCup:Sip
  HCo    ~ 1 * HCoCup:Sip
"
cost  <- costF(actions, cost = -0.1)

field <- labelF(paste(afford, cost,
                      "L ~ 0.9 * HCo + 0.1 * Energy", sep = "\n"),
                actions = actions)
field$params["use"] <- -1          # consumption strength (the shared label)

cat(field$model)
field$params
```

[`labelF()`](../reference/labelF.html) gives every fixed force its
canonical `fZ_X.Y` label and captures its value; the one parameter it
leaves for us is the shared **`use`** label, the consumption strength,
which we set to `-1`. Note what is *not* here: no `make`, `TurnOn`, `Pour`
or `Sip` ever appears on a left-hand side. **Actions are exogenous** — the
field says what each action *would* do; it is the policy, next, that
decides which one actually fires.

## Step 2: The policy

A **policy** is a set of `action ~ condition` rules: fire each action once
its prerequisite state is present. The first action is launched not by a
situational state but by the agent's **`choice`** to act — a one-shot
intention that is spent the moment it sets things in motion.

```{r policies}
full <- "
  make   ~ 1 * choice
  TurnOn ~ 1 * s2
  Pour   ~ 1 * HCoPot
  Sip    ~ 1 * HCoCup
"
```

To have something to compare against later, we also draft a **faulty**
plan that skips a step — turning the machine on and reaching for a sip
without ever pouring:

```{r policies-faulty}
faulty <- "
  make   ~ 1 * choice
  TurnOn ~ 1 * s2
  Sip    ~ 1 * HCoCup
"
```

`simulateF()` reads only the *condition variables* of each rule (the
coefficients are ignored for eligibility — they are there so the same
string can be drawn as edges later). A rule says *when an action is
allowed*, not how strong it is.

## Step 3: Run the field

The agent starts committed (`choice = 1`), everything else off:

```{r s0}
s_0 <- c(choice = 1, make = 0, s2 = 0, TurnOn = 0, HCoPot = 0,
         Pour = 0, HCoCup = 0, Sip = 0, HCo = 0, Energy = 0, L = 0)

out <- simulateF(field$model, field$params, s_0, full, flows = "choice")
round(out$trajectory[, c("choice","make","s2","TurnOn","HCoPot",
                         "Pour","HCoCup","Sip","HCo","Energy","L")], 2)
out$fired
```

One stage per turn, and **each action fires exactly once**. Read a turn
left to right:

- **eligibility**: the actions whose condition currently holds. At the
  start only `make` qualifies (its trigger `choice` is on); the rest are
  gated on states that do not yet exist.
- **fire one**: of the eligible actions, one fires. Here only one ever
  qualifies at a time, so the choice is forced; when several qualify,
  `simulateF()` *appraises* them — simulates each one turn ahead and fires
  the one that most raises `L` (the deductive job of a choice node).
- **settle**: the field takes one step — the fired action produces the
  next state and consumes the one that afforded it, and `Energy` takes its
  `-0.1`. `flows = "choice"` marks `choice` a one-shot, spent once `make`
  reads it.

Consumption is what keeps the run moving and finite: once `TurnOn`
consumes `s2`, `TurnOn` is no longer eligible, so it cannot re-fire; the
token simply hands itself down the line. By the last turn `HCo = 1`,
`Energy = -0.4` (four actions paid for), and the values vector nets these
to `L = .9 \cdot 1 + .1 \cdot (-0.4) = 0.86`. The run ends because nothing
is eligible any more:

```{r halted}
out$halted
```

## Step 4: Choosing among plans

The "which plan?" question needs no separate machinery: **run each policy
and compare the final value.** That comparison *is* the choice node, one
level up from the per-turn appraisal — an argmax over what each plan
affords, net of what it costs. [`chooseF()`](../reference/chooseF.html)
wraps it: run each policy to quiescence, tabulate the final state, and flag
the winner. It appends a do-nothing baseline by default.

```{r choose}
chooseF(field$model, field$params, s_0,
        policies = list(full = full, faulty = faulty), flows = "choice")
```

The full plan reaches coffee and, net of cost, earns `L = 0.86`; doing
nothing reaches nothing, spends nothing, and earns `L = 0`. The telling
row is **`faulty`**. It makes and turns on — two real actions, so it has
spent `Energy = -0.2` — but never pours, so the cup is never filled, the
`Sip → HCo` force stays gated shut, and `HCo` stays `0`. With nothing on
the value side to show for it, its utility goes **negative**
(`L = -0.02`): the faulty plan scores *below doing nothing*. That is the
payoff of accounting for cost — without the energy term the two would tie
at `L = 0`; with it they come apart, and inaction wins by conserving
effort. Causal gating decides whether a plan reaches the outcome; the cost
ledger decides what a *failed* reach is worth.

## Step 5: Watch it unfold

[`plotField()`](../reference/plotField.html) draws the field at one state
as a node-link diagram: the **choice point** is a circle, **object**
states are diamonds, **actions** right-facing triangles, the **value**
readout a left-facing triangle, and **`Energy`** a square holding a ⚡
bolt, set just left of `L` so the value forces from `HCo` (down the
staircase) and `Energy` (along the floor) converge on `L`. Every node is
shaded on a diverging **red–white–blue** scale (negative red, zero white,
positive blue). Each edge encodes its force, with the level printed on the
shaft: thicker and darker for larger `|f|`, dashed when negative — so the
production forces (`f = 1`) read as strong solid lines and the energy
costs (`f = -0.1`) as faint dashed ones. The **policy** edges are drawn in
**gold**, distinct from the field's **black**; the backward **consumption**
arcs bow below their forward edge so they do not overlay it.

We seat `L` and `Energy` on the floor, level with `make`, so the diagram
closes back where it began. For drawing, we layer the policy onto the
field so its gold intention-edges appear:

```{r layout}
model_full <- paste(field$model, full, sep = "\n")

layout <- data.frame(
  name  = c("choice","make","s2","TurnOn","HCoPot",
            "Pour","HCoCup","Sip","HCo","Energy","L"),
  x     = c(0,    2,     3,   5,       6,
            8,     9,      11,   12,   12,   14),
  y     = c(0,    0,     2,   2,       4,
            4,     6,      6,    8,    0,    0),
  shape = c("circle","rtTri","diamond","rtTri","diamond",
            "rtTri","diamond","rtTri","diamond","square","lfTri"),
  label = c("choice","make","rig\nset","turn on","coffee\nin pot",
            "pour","coffee\nin cup","sip","have\ncoffee","Energy","L"),
  glyph = c(NA, NA, NA, NA, NA, NA, NA, NA, NA, "⚡", NA),
  stringsAsFactors = FALSE
)
```

Stepping `plotField()` across the trajectory and stitching the frames
with [`plotsAsWidget()`](../reference/plotsAsWidget.html) animates the
run. Use **← Back** / **Forward →** to watch the choice point flash and
empty, then a single lit node climb the staircase — each action firing as
its stage is reached, the state it produced lighting beside it, and the
previous stage emptying as it is consumed — while the ⚡ square logs the
cost:

```{r field-widget, eval=has_ggplot, fig.alt="Interactive staircase field diagram stepping through the coffee simulation; the choice point fires once at the foot of the staircase, then a single lit node climbs as each action fires and each stage is consumed."}
steps  <- nrow(out$trajectory) - 1L
frames <- lapply(0:steps, function(t) {
  plotField(model_full, field$params, out$trajectory[paste0("t=", t), ],
            layout, plan = full, title = paste("t =", t))
})

plotsAsWidget(
  frames,
  panel_titles = paste("Step", 0:steps),
  width = 9, height = 6
)
```

## Step 6: Where a faulty plan breaks

Running the faulty policy through the *same* field shows **where** it
breaks. It commits to making, turning on, and (eventually) sipping, but
never pours:

```{r faulty-run}
out_np <- simulateF(field$model, field$params, s_0, faulty, flows = "choice")
round(out_np$trajectory[, c("make","s2","TurnOn","HCoPot","Pour",
                            "HCoCup","Sip","HCo","Energy","L")], 2)
out_np$fired
```

The token climbs two stages — `make` sets up the rig, `TurnOn` brews
`HCoPot` — and then **stops**. With no pour rule, `Pour` is never
eligible, so the coffee sits in the pot, the cup is never filled, and the
run goes quiescent with `HCo` never reached. The two actions taken have
already cost `Energy = -0.2`, so the readout settles **negative**
(`L = -0.02`). Stepping through, the lit node climbs to **coffee in pot**
and halts — the staircase ends mid-flight:

```{r faulty-widget, eval=has_ggplot, fig.alt="Interactive staircase field diagram for the faulty (no-pour) plan; the lit token climbs to coffee-in-pot and halts there, with the pour, cup, sip and have-coffee stages left dark."}
model_np  <- paste(field$model, faulty, sep = "\n")
steps_np  <- nrow(out_np$trajectory) - 1L
frames_np <- lapply(0:steps_np, function(t) {
  plotField(model_np, field$params, out_np$trajectory[paste0("t=", t), ],
            layout, plan = faulty, title = paste("t =", t))
})

plotsAsWidget(
  frames_np,
  panel_titles = paste("Step", 0:steps_np),
  width = 9, height = 6
)
```

Tellingly, the pour *affordance* stays open: because there is coffee in
the pot, the `Pour → coffee-in-cup` force (and its consumption arc back to
the pot) are still drawn — but they leave a **dark `pour` node**. The move
is available and the field would reward it; the policy simply never calls
it. That gap is the whole failure: one missing rule, and everything
downstream of it stays unlit. The run doesn't error or stall on a
technicality — it **runs out of eligible actions**, which is exactly what
a broken plan is.

## Step 7: Blowing out the prep — parallel, any-order actions

So far every action had exactly one predecessor: the pipeline self-ordered
through gating. Real plans are rarely so linear. The single `make` step
compressed all of *setting up the coffee maker* into one move; blow it out
into the five things that actually have to happen — add water, add grounds,
add a filter, plug it in, set the pot under it — and two new features of
action appear:

- the five preps can be done in **any order** (they don't depend on each
  other), and
- the machine brews only once **all five** are done (a **conjunction**).

Each prep is an action producing its own state. "Prepped" is the AND of all
five — and since `lavaan` interactions are only pairwise,
[`andF()`](../reference/andF.html) builds that five-way conjunction two at a
time, returning the model fragment and the names of the **derived** nodes
it introduces:

```{r blowout-field}
preps    <- c("AddWater", "AddGrounds", "AddFilter", "PlugIn", "AddPot")
haves    <- c("HasWater", "HasGrounds", "HasFilter", "HasPower", "HasPot")
actions2 <- c(preps, "TurnOnCM", "Pour", "Sip")

prepped <- andF("Prepped", haves)
cat(prepped$model)
prepped$aux

field2 <- labelF(paste("
  HasWater   ~ 1 * AddWater + use * HasWater:TurnOnCM
  HasGrounds ~ 1 * AddGrounds
  HasFilter  ~ 1 * AddFilter
  HasPower   ~ 1 * PlugIn
  HasPot     ~ 1 * AddPot",
  prepped$model,
  "HCoPot ~ 1 * Prepped:TurnOnCM + use * HCoPot:Pour
   HCoCup ~ 1 * HCoPot:Pour      + use * HCoCup:Sip
   HCo    ~ 1 * HCoCup:Sip",
  costF(actions2, cost = -0.1),
  "L ~ 0.9 * HCo + 0.1 * Energy", sep = "\n"),
  actions = actions2)
field2$params["use"] <- -1
```

Brewing (`TurnOnCM`) is gated on `Prepped` and *uses up the water*, which
un-prepares the machine so it can't re-brew. The agent's plan is a **to-do
list**: five one-shot intentions, one per prep, plus the usual brew / pour /
sip rules. Because all five intentions are live at the start, all five preps
are eligible at once:

```{r blowout-policy}
policy2 <- "
  AddWater   ~ toWater
  AddGrounds ~ toGrounds
  AddFilter  ~ toFilter
  PlugIn     ~ toPower
  AddPot     ~ toPot
  TurnOnCM   ~ Prepped
  Pour       ~ HCoPot
  Sip        ~ HCoCup
"
todo <- c(toWater = 1, toGrounds = 1, toFilter = 1, toPower = 1, toPot = 1)
s2_0 <- c(todo,
          setNames(rep(0, length(haves)), haves),
          setNames(rep(0, length(prepped$aux)), prepped$aux),
          setNames(rep(0, length(actions2)), actions2),
          HCoPot = 0, HCoCup = 0, HCo = 0, Energy = 0, L = 0)
```

Run it. `aux = prepped$aux` tells `simulateF()` those conjunction nodes are
**derived** — recomputed each turn from the current states rather than
latching — and `flows` marks the five to-do tokens as one-shot:

```{r blowout-run}
out2 <- simulateF(field2$model, field2$params, s2_0, policy2,
                  aux = prepped$aux, flows = names(todo), max_turns = 20)
round(out2$trajectory[, c("HasWater","HasGrounds","HasFilter","HasPower",
                          "HasPot","Prepped","HCoPot","HCoCup","HCo","Energy","L")], 2)
out2$fired
```

The five preps fire one per turn (turns 1–5), `Prepped` flips on the moment
the last one lands, brewing fills the pot at turn 6, and pour / sip finish
the job — eight actions, so `Energy = -0.8` and `L = 0.82`. Two things are
worth drawing out:

- **The order is genuinely arbitrary.** When several actions are eligible,
  the choice node appraises each one turn ahead and takes the best — but no
  single prep moves `L` (you need all five *and* the brew), so the appraisal
  is **indifferent**, and the tie is broken by listing order. Any permutation
  of the five reaches the same prepped machine and the same final `L`. This
  is the first time the per-turn appraisal faces a real tie; the linear
  pipeline never did.
- **`Prepped` is a derived node, not a stock.** It is a pure function of the
  five states, so it recomputes every turn (passed via `aux`) and
  `simulateF()` resolves the whole conjunction chain within a single turn.
  The moment water is consumed at brew time, `Prepped` falls back to `0`.

Drawn, the process is a **fan into a staircase**: five parallel preps merge
into `Prepped`, then the familiar linear brew → pour → sip. The merge edges
are **conjunctive** --- `Prepped` fires only when *all* five hold --- so
`plotField()` draws them **dotted** (a product), rather than the solid
numeric edges of an additive relationship, which would misleadingly read as
regression weights. (For the diagram we show the
conjunction as a direct five-way merge; under the hood it is the pairwise
chain `andF()` built.)

```{r blowout-widget, eval=has_ggplot, fig.alt="Interactive field diagram for the blown-out prep: five prep actions in parallel converge on a 'prepped' node, then a linear brew-pour-sip staircase to coffee."}
disp2 <- "
  HasWater ~ 1*AddWater
  HasGrounds ~ 1*AddGrounds
  HasFilter ~ 1*AddFilter
  HasPower ~ 1*PlugIn
  HasPot ~ 1*AddPot
  Prepped ~ 1*HasWater + 1*HasGrounds + 1*HasFilter + 1*HasPower + 1*HasPot
  HCoPot ~ 1*Prepped:TurnOnCM
  HCoCup ~ 1*HCoPot:Pour
  HCo ~ 1*HCoCup:Sip
  L ~ 0.9*HCo + 0.1*Energy"
disp_pol2   <- "TurnOnCM ~ 1*Prepped\nPour ~ 1*HCoPot\nSip ~ 1*HCoCup"
disp_model2 <- paste(disp2, disp_pol2, sep = "\n")

layout2 <- data.frame(
  name  = c(preps, haves, "Prepped","TurnOnCM","HCoPot","Pour",
            "HCoCup","Sip","HCo","Energy","L"),
  x     = c(rep(0,5), rep(2,5), 4.5, 6.0, 7.3, 8.6, 9.9, 11.2, 12.5, 10.0, 14.0),
  y     = c(0,2,4,6,8, 0,2,4,6,8, 4, 5.5, 4, 5.5, 4, 5.5, 4, 0.6, 4),
  shape = c(rep("rtTri",5), rep("diamond",5), "diamond","rtTri","diamond",
            "rtTri","diamond","rtTri","diamond","square","lfTri"),
  label = c("add\nwater","add\ngrounds","add\nfilter","plug\nin","add\npot",
            "water","grounds","filter","power","pot","prepped","turn\non",
            "pot\ncoffee","pour","cup\ncoffee","sip","have\ncoffee","Energy","L"),
  glyph = c(rep(NA,17), "⚡", NA),
  stringsAsFactors = FALSE)

steps2  <- nrow(out2$trajectory) - 1L
frames2 <- lapply(0:steps2, function(t) {
  plotField(disp_model2, numeric(0), out2$trajectory[paste0("t=", t), ],
            layout2, plan = disp_pol2, conjunctive = "Prepped",
            node_size = 0.30, title = paste("t =", t))
})

plotsAsWidget(frames2, panel_titles = paste("Step", 0:steps2),
              width = 10, height = 6)
```

## Consumption is what keeps a field bounded

`simulateF()` does **not** clamp. Functional-field state is expected to
stay within `[-1, 1]`; a value outside it signals a misspecification
rather than something to silently truncate. The clearest way to trip it is
to give a stock an inflow but no consumption — and a policy that keeps
firing the action that feeds it. With nothing to use the stock up, it
accumulates without bound and the run never goes quiescent:

```{r runaway, warning=TRUE}
runaway <- simulateF("tank ~ 1 * pump", params = numeric(0),
                     s_0 = c(on = 1, pump = 0, tank = 0),
                     policy = "pump ~ on", max_turns = 4)
runaway$trajectory[, "tank"]
```

The warning names the offending node and turn, and the true (un-clamped)
value is returned so the overshoot is visible. In the coffee field the
`use` terms are exactly what keep every stock bounded: each stage hands
its token on and empties, so nothing piles up — and the run *ends*,
because once everything is consumed, nothing is eligible.

## Takeaway

One **field**, one **policy**, one **run**. The field is the physics — how
actions produce and consume states, what they cost, and what the agent
values; the policy is the plan — which action is allowed when; and
[`simulateF()`](../reference/simulateF.html) plays the policy on the field
one action at a time. Because actions enter singly and exogenously, the
same run is both the *equilibrium read* (its final `L` is what the plan
affords) and the *temporal read* (its trajectory is how the plan unfolds)
— there is no separate mode for the two questions.

The division of labour stays explicit throughout: the agent's **values**
are a property of the field (the vector `f[., L]`, here `.9` on having
coffee and `.1` on keeping energy); the **policy** (gold) holds the
standing intentions; the field's **production** decides whether a fired
action accomplishes anything, its **consumption** hands each stage on and
keeps the run finite, and [`costF()`](../reference/costF.html)'s decrement
is charged for every action *attempted*. Accounting for cost is what lets
a plan score *below doing nothing*: the skip-the-pour plan affords no
coffee yet still spends energy, so it is worse than inaction — and a
choice node, comparing final values, never picks it.

The same machinery generalises beyond one agent: a turn need not always
belong to the same actor. Let agent **A** act, the field re-settle, and
agent **B** appraise the *changed* field and respond — and you have a
two-party interaction in the same engine. That is the subject of a later
vignette. The inductive direction — recovering the field's coefficients
from data — is the subject of the `pathXMY` vignettes
(`vignette("speeding_pathXMY", package = "funfield")` and its companions).
```
