---
title: "Coffee Field Model: Sequential Action Planning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Coffee Field Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4
)
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
```

## What is a functional field?

A **functional field** is a model of how situational forces propagate
through an environment to produce action outcomes. At each moment the
field specifies which forces are *afforded* — which causal pathways are
open given the current state of the situation. Forces spread
step-by-step: a disturbance at time *t* is transmitted through the field
to produce new state at time *t + 1*, which then opens (or closes) the
pathways available at *t + 2*.

The same field `F` supports two complementary workflows. **Induction**
runs data → `F`: you fit the field's coefficients from observed
behaviour with `pathXMY()` and friends. **Deduction** runs `F` →
behaviour: you treat the coefficients as a working theory and *simulate*
forward to see what a plan produces. This vignette is about the second
direction. For the notation it uses throughout — the `F` / `F_t` field
objects and the `fZ_X.Y` parameter schema — see
`vignette("notation", package = "funfield")`.

The worked example is making a cup of drip coffee: a four-step action
sequence that becomes causally connected through a field. The key idea
is **causal gating** — a later step only has an effect once an earlier
step has changed the situation. The plan supplies the *intention* to act
at each stage, but the situation controls whether acting does anything.

## The Markov step

A functional field iterates

\[
F_t = F(s_t), \qquad s_{t+1} = s_t \cdot F_t,
\]

where `s_t` is the situational **state** (a named vector of node values)
and `F_t` is the field evaluated at that state. Two helpers run this
loop:

- [`evalF()`](../reference/evalF.html) takes one step: given the field
  (as a `lavaan` model string), its parameter values, and the current
  `s_t`, it returns `s_{t+1}`.
- [`runF()`](../reference/runF.html) drives the iteration for a fixed
  number of steps and returns the full trajectory. (It can also inject
  externally scheduled **actions**, but here we let the plan drive
  itself — see Step 2.)

We build the field in two halves that `lavaan` lets us stitch together
by simple concatenation: the **situation** — the affordance structure of
the world, which exists whether or not anyone acts — and the **action
plan** — a set of conditional intentions the agent layers on top.

## Step 1: The situation

Each stage of the coffee process is an **afforded force**: a path that
contributes nothing until its prerequisite state variable is on. In the
`fZ_X.Y` schema an afforded force is a term with a moderator but no
baseline — `fZ_X.Y * X:Z` — so the `X → Y` force is zero whenever
`s[Z] = 0` and switches on once `s[Z]` flips positive.

The four gated stages, plus a terminal utility force:

| Force | Afforded by | Reads as |
|---|---|---|
| `make → s2`        | `s1`     | setting up the rig works once you're ready to make coffee |
| `TurnOn → HCoPot`  | `s2`     | turning it on brews into the pot once the rig is ready |
| `Pour → HCoCup`    | `HCoPot` | pouring fills the cup once there's coffee in the pot |
| `Sip → HCo`        | `HCoCup` | sipping yields "having coffee" once it's in the cup |
| `HCo → L`          | —        | having coffee produces likelihood / value (baseline) |

The terminal node is **`L`** — likelihood. In ESJT data `L` is the rated
likelihood of taking an action, and here it plays the role an economist
would give *subjective expected utility*: the field assumes a person
takes the action they appraise most positively — the one they see, at
some level, as carrying the highest utility. Reading `L` off the
settled field is therefore reading off which action the agent is
predicted to choose.

Node names are lavaan-safe (letters, digits, `_` only); the `.` inside
each parameter label is the source/target separator. We need not type
those labels by hand: [`labelF()`](../reference/labelF.html) derives the
`fZ_X.Y` label for each term and captures any coefficient we fix inline
into a matching `params` vector. The labelling is **order-dependent** and
reads straight off the right-hand side — `Y ~ Z:X` becomes `fZ_X.Y` — so
each interaction is written **condition first, action last**: the gating
state, then the action it affords. We fix a force of `1` everywhere, so
the wave propagates without attenuation (smaller values would damp it as
it travels down the chain). Passing `actions =` lets `labelF()` check
that no action has slipped into a condition slot:

```{r situation}
library(funfield)

situation <- labelF("
  s2     ~ 1 * s1:make
  HCoPot ~ 1 * s2:TurnOn
  HCoCup ~ 1 * HCoPot:Pour
  HCo    ~ 1 * HCoCup:Sip
  L      ~ 1 * HCo
", actions = c("make", "TurnOn", "Pour", "Sip"))

cat(situation$model)
situation$params
```

`labelF()` returns the relabelled `model` string and its `params`. Each
interaction reads condition → action: `s2:TurnOn` is the force from the
action `TurnOn` to `HCoPot`, afforded by the state `s2`, and labelled
`fs2_TurnOn.HCoPot`. The lone baseline term `L ~ HCo` has no condition
and becomes `f1_HCo.L` — always active once `HCo` is present. Nothing
here actually *fires* an action — the situation says only what each
action *would* produce, given the right prerequisites. The firing is the
plan's job.

## Step 2: The action plan

A **conditional action plan** says *when* to act: fire each action once
its prerequisite situational state is achieved — "set up the rig once
I'm ready, turn it on once the rig is set, pour once it's brewed, sip
once it's poured." Each rule is a force from a *state* to an *action*,
which we run through `labelF()` the same way:

```{r plan}
plan <- labelF("
  make   ~ 1 * s1
  TurnOn ~ 1 * s2
  Pour   ~ 1 * HCoPot
  Sip    ~ 1 * HCoCup
")

cat(plan$model)
```

Two things to notice. First, these forces point the *opposite* way from
the situation's: the situation runs action → outcome, the plan runs
state → action. Second, fixing each trigger at `1` is a structural
commitment — "act, fully, once the prerequisite holds" — rather than a
force to be estimated from data; what distinguishes a plan force from a
situation force in what follows is not its value but its colour (gold,
below).

Because both halves are plain `lavaan` model strings, we stitch the
field together — joining with a newline so the two blocks stay on
separate lines — and combine their parameter vectors:

```{r stitch}
model  <- paste(situation$model, plan$model, sep = "\n")
params <- c(situation$params, plan$params)
```

The plan replaces the externally scheduled `actions` list used in
earlier versions of this vignette: instead of telling the simulation
*at which step* to act, we tell it *under what conditions* to act, and
let the dynamics work out the timing.

## Step 3: Run the simulation

The state starts with the agent **ready to make coffee** (`s1 = 1`) and
everything else off. With the plan now inside the field, `runF()` needs
no `actions` argument — it just propagates. Each stage now takes *two*
hops (state → action, then action → outcome), so the chain needs about
twice as many steps to settle as the externally-scheduled version did.

```{r run}
s_0 <- c(s1 = 1, make = 0, s2 = 0, TurnOn = 0, HCoPot = 0,
         Pour = 0, HCoCup = 0, Sip = 0, HCo = 0, L = 0)

traj <- runF(model, params, s_0, steps = 10)
round(traj, 2)
```

Read the trajectory as a wave climbing one node per step, alternating
between an action firing and the outcome it affords:

| Step | What fires | Driven by |
|---|---|---|
| 1 | `make`   | plan: ready (`s1`) → make |
| 2 | `s2`     | situation: make → rig ready |
| 3 | `TurnOn` | plan: rig ready (`s2`) → turn on |
| 4 | `HCoPot` | situation: turn on → coffee in pot |
| … | …        | … |
| 9 | `L`      | situation: have coffee → likelihood |

The final state is identical to the scheduled version's — every node
ends at `1` — but it is reached by the plan triggering itself off the
states the situation produces, rather than by an external schedule. The
gating still does the real work: `TurnOn`'s plan rule is keyed on `s2`,
so even though the intention to turn on is present from the start, it
fires nothing until the rig is actually set up.

## Step 4: Visualise the field as a staircase

`plotField()` draws the field at a single state as a node-link diagram.
Following the field-diagram convention, **object** nodes (the situational
states) are diamonds, **action** nodes are right-facing triangles, and
the terminal **likelihood** node is a left-facing triangle. Each node is
shaded by its current value — white at `0`, blue at `1` — and an arrow is
drawn for every force that is currently *afforded*: the resolved field
coefficient `params[label] * s[gate]` collapses an afforded force
`1 · X:Z` to `0` (gate shut, no edge) or `1` (gate open, edge present)
once the level of the gating state `Z` is known.

Passing `plan = plan` colours the forces the action plan contributes in
**gold**, distinct from the situation's own forces in grey — a standing
convention for "forces added by a conditional action plan." Laid out as a
staircase, the two interleave: gold edges are the *horizontal* steps
(a state handing off to the action it triggers) and grey edges are the
*diagonal* steps (an action producing its outcome).

The node spacing is chosen so the two kinds of arrow stay close in
length and the staircase climbs steeply enough to fill a roughly square
frame rather than a long thin strip: gold state → action steps span two
horizontal units, and grey action → outcome steps span one horizontal
and two vertical units:

```{r layout}
layout <- data.frame(
  name  = c("s1","make","s2","TurnOn","HCoPot",
            "Pour","HCoCup","Sip","HCo","L"),
  x     = c(0,    2,     3,   5,       6,
            8,     9,      11,   12,   14),
  y     = c(0,    0,     2,   2,       4,
            4,     6,      6,    8,    8),
  shape = c("diamond","rtTri","diamond","rtTri","diamond",
            "rtTri","diamond","rtTri","diamond","lfTri"),
  label = c("ready","make","rig\nset","turn on","coffee\nin pot",
            "pour","coffee\nin cup","sip","have\ncoffee","L"),
  stringsAsFactors = FALSE
)
```

Stepping `plotField()` across the rows of the trajectory and stitching
the frames with `plotsAsWidget()` gives an interactive back/forward
animation. Use **← Back** / **Forward →** to watch forces appear as each
outcome is attained and nodes turn blue as state propagates up the
staircase:

```{r field-widget, eval=has_ggplot, fig.alt="Interactive staircase field diagram stepping through the coffee simulation."}
frames <- lapply(0:10, function(t) {
  plotField(model, params, traj[paste0("t=", t), ], layout,
            plan = plan$model, title = paste("t =", t))
})

plotsAsWidget(
  frames,
  panel_titles = paste("Step", 0:10),
  width = 9, height = 6
)
```

At `t = 0` the entire plan is already laid out in gold — the agent holds
all four conditional intentions from the start — but only `make → rig
set` (afforded because the agent begins *ready*, `s1 = 1`) and the
baseline `have coffee → L` are live among the grey situation forces. As
the wave climbs, each newly-attained object opens the next situation
force, a fresh grey arrow appears one stage up, and nodes turn blue as
the action they gate finally fires.

### The same wave as a heatmap

The diagonal structure also reads cleanly as a stage × time tile plot — a
compact way to see all stages and all steps at once:

```{r wave, eval=has_ggplot, fig.alt="Tile plot of state value by process stage and time step, showing a diagonal wave."}
library(ggplot2)

stage_order  <- layout$name
stage_labels <- setNames(gsub("\n", " ", layout$label), layout$name)

long <- data.frame(
  step  = rep(0:10, times = length(stage_order)),
  stage = factor(rep(stage_order, each = nrow(traj)),
                 levels = rev(stage_order)),
  value = as.vector(traj[, stage_order])
)

ggplot(long, aes(step, stage, fill = value)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  scale_fill_gradient(low = "white", high = "#1572da", limits = c(0, 1)) +
  scale_x_continuous(breaks = 0:10, expand = c(0, 0)) +
  scale_y_discrete(labels = stage_labels) +
  labs(x = "time step", y = NULL, fill = "state",
       title = "The coffee field's wave of afforded forces") +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank(),
        legend.position = "right")
```

## Takeaway

Sipping is not merely *last* in time — it is only *causally possible*
once coffee is in the cup, which requires having poured, which requires
having brewed, which requires setting up the machine. Splitting the field
into a **situation** and a **conditional action plan** makes the division
of labour explicit: the plan (gold) holds standing intentions keyed on
states, but the situation's **afforded forces** (grey) decide whether a
fired action produces anything. The agent can intend to turn the machine
on from the very first step, yet nothing brews until the rig is actually
set up — the gate keyed on `s2` stays shut until then.

This is the deductive use of a functional field: given a theory of the
world (the situation's `fZ_X.Y` coefficients) and a plan (the conditional
triggers), simulate forward and read off what the plan produces. The
inductive direction — recovering those situation coefficients from data —
is the subject of the `pathXMY` vignettes
(`vignette("speeding_pathXMY", package = "funfield")` and its
companions).
