---
title: "Moderated mediation with `pathXMY()`: the coworker assertiveness ESJT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Moderated mediation with pathXMY(): the coworker assertiveness ESJT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png",
  dpi = 192,
  out.width = "70%"
)
```

This vignette demonstrates moderated mediation with `pathXMY()` using the
`coworkerESJT` dataset — a workplace assertiveness ESJT in which participants
rated two possible responses across 12 scenarios involving a coworker making an
inappropriate request. Three moderators are examined: the **rank** of the
coworker in the scenario (situation-level; experimentally assigned), the
**perceived unethicality** of the coworker's request (situation-level;
participant-rated), and **boldness** as a self-reported personality trait
(person-level).

## The data

```{r data}
library(funfield)
data(coworkerESJT)
lapply(coworkerESJT, dim)
```

The dataset is a named list with five components:

- **`$PSI`** — 5,846 rows: 249 persons × 12 situations × 2 actions, with 8
  rated outcome features and `Likelihood`
- **`$cond`** — situational condition per (person, scenario): `subordinate`,
  `coworker`, or `manager` (randomly assigned, ~4 per condition per person)
- **`$sit`** — situation-level perceptions (action-independent; ~82.5%
  coverage)
- **`$IIDL`** — 40 IIDL personality items for 206 of the 249 participants
- **`$codebook`** — full survey wording, scale anchors, and scenario texts

## Variables and survey wording

The predictor, mediators, and outcome are the expected-outcome ratings from
`$PSI`. Participants rated each action on the stem:

> *"If you do this... [action] ...how likely is this outcome to result?"*

The table below shows the variable name and the complete survey wording for
each item (−1 = Very Unlikely, 1 = Very Likely):

```{r wording-table}
knitr::kable(
  coworkerESJT$codebook$outcomes[, c("var", "label")],
  col.names = c("Variable", "Survey wording"),
  caption = "Outcome items (PSI)"
)
```

The focal **predictor** is `ExpDisapproval` — *"Having directly/forcefully
expressed disapproval with your [coworker]'s behavior."* Note that this is
itself an expected outcome: respondents rate how likely it is that taking this
action will result in them having expressed disapproval. It is coded such that
higher values mean the action is *more* likely to constitute such an
expression.

The focal **outcome** is `Likelihood` — *"How likely are you to do the
following actions in response to this situation?"*

The seven **mediators** are the remaining expected-outcome features.

## Data preparation

Merge the condition variable into the PSI data, code a `manager` dummy
(1 = coworker is a manager, 0 = coworker or subordinate), then within-person
deviate all Level-1 columns.

```{r prep}
L1cols <- c("Appropriate","ExpDisapproval","MajorArg","GoodRelationP",
            "Punished","CompleteJob","P_Punished","CompanySuccess","Likelihood")
mediators <- c("Appropriate","MajorArg","GoodRelationP",
               "Punished","CompleteJob","P_Punished","CompanySuccess")

dat <- merge(coworkerESJT$PSI, coworkerESJT$cond, by = c("p","s"))
dat$manager <- as.numeric(dat$condition == "manager")

for (v in L1cols) {
  dat[[v]] <- ave(dat[[v]], dat$p, FUN = function(x) x - mean(x))
}
```

## Normative model (no moderator)

How does `ExpDisapproval` — the expectation of having directly expressed
disapproval — mediate the link between an action and how likely the respondent
is to take it, averaging across people and conditions?

```{r normative}
norm <- pathXMY(dat, X = "ExpDisapproval", Y = "Likelihood", M = mediators)

## Loop pass: one X -> M -> Y regression PER MEDIATOR, fit independently.
## Each B1_YM row is the slope of Likelihood on that mediator controlling
## only for ExpDisapproval (no other mediators in the equation).
norm$tidy_loop

## Joint pass: a SINGLE simultaneous regression with all mediators together
## in the Y equation. B1_YM_joint slopes are partial slopes net of the
## other mediators -- not comparable to the loop B1_YM above.
norm$tidy_joint
```

### Expectation paths (B1_MX): what co-occurs with expressing disapproval?

Actions that score higher on `ExpDisapproval` (i.e., actions that would
constitute more direct expression of disapproval) are also seen as more likely
to produce a major argument/confrontation (`MajorArg`, B1_MX ≈ 0.70) and less
likely to maintain a good working relationship in the long run (`GoodRelationP`,
B1_MX ≈ −0.52). They are also seen as more appropriate within the role
(`Appropriate`, B1_MX > 0) and more likely to result in the coworker being
formally punished (`P_Punished`, B1_MX > 0).

### Valuation paths (B1_YM): which outcomes drive action likelihood?

Acting appropriately (`Appropriate`, B1_YM ≈ 0.73) and benefiting company
success (`CompanySuccess`, B1_YM ≈ 0.74) are the strongest positive drivers of
whether a respondent says they would take an action. Being formally punished
(`Punished`, B1_YM ≈ −0.56) and having a major argument (`MajorArg`,
B1_YM ≈ −0.34) are the strongest suppressors.

### Indirect effects (B1_MX × B1_YM)

Indirect effects are significant for all mediators where both component paths
are non-zero. The indirect path through `Appropriate` is particularly large:
actions that more strongly constitute an expression of disapproval are also
seen as more role-appropriate, and perceived appropriateness strongly drives
action likelihood.

## Situation-level moderation: coworker rank

Does the rank of the coworker in the scenario change how expressing disapproval
co-occurs with other outcome expectations? The `manager` dummy varies *within*
persons across scenarios (randomly assigned), so we use `Z.within = TRUE`.
`pathXMY()` then within-person deviates the moderator before fitting.

```{r mod-situation}
mod_s <- pathXMY(dat, X = "ExpDisapproval", Y = "Likelihood",
                 M = mediators, Z = "manager", Z.within = TRUE)
```

### Does manager rank moderate the expectation arm (ExpDisapproval → M)?

```{r bz-mx}
knitr::kable(
  subset(mod_s$tidy_loop, param == "BZ_MX")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Moderation of expectation paths by manager rank (BZ_MX)"
)
```

Six of the seven mediators show significant moderation. When the coworker holds
a higher rank (manager condition), expressing disapproval is expected to be
substantially *more* likely to result in formal punishment of oneself
(`Punished`, BZ_MX > 0): the power asymmetry increases the perceived personal
cost. At the same time, expressing disapproval in a manager scenario is
perceived as *less* likely to be seen as role-appropriate (`Appropriate`), less
likely to help complete the job (`CompleteJob`), less likely to maintain a good
working relationship (`GoodRelationP`), and less likely to benefit company
success (`CompanySuccess`) — the perceived benefits of asserting shrink when
facing a higher-power colleague.

### Does that moderation carry through to the indirect effect?

```{r bz-indirect}
knitr::kable(
  subset(mod_s$tidy_loop, param == "BZ_MX * B1_YM")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Moderation of indirect effects by manager rank (BZ_MX × B1_YM)"
)
```

All significant expectation moderation translates into significant moderation
of the indirect effect. The indirect path through `Punished` is the largest in
magnitude: in manager scenarios, the combination of expecting more formal
punishment *and* punishment's strong negative valuation substantially suppresses
the indirect pull toward expressing disapproval through that path.

### Decomposition and field view

`pathXMY_decompose()` partitions the total Z-moderation (`BZ_YX[1]`)
into expectation routing (`BZ_MX × B1_YM`), valuation routing
(`B1_MX × BZ_YM`), and a residual direct term. The static
`plotPathXMY()` shows the full field with `b1 + bZ(Z)` edge labels;
the toggleable widget overlays the field at low / mid / high
manager exposure.

```{r decompose-manager}
dec_m <- pathXMY_decompose(dat, X = "ExpDisapproval", Y = "Likelihood",
                           M = mediators, Z = "manager", Z.within = TRUE)
knitr::kable(dec_m$total, digits = 3,
             caption = "Total moderation BZ_YX[1] for manager rank")
```

```{r plot-manager, fig.width = 10, fig.height = 7}
plotPathXMY(dec_m,
            X_label = "ExpDisapp", Y_label = "Likelihood",
            Z_label = "manager",
            title = "ExpDisapproval -> M -> Likelihood (Z = manager)")
```

The fan diagram's mediator arms are the per-mediator *loop* fits
(one X to M to Y model per mediator, overlaid). The straight
**X to Likelihood arrow** is the residual direct path from the
*joint* multi-mediator fit (added automatically by
`pathXMY(joint = TRUE)`, the default) — i.e. the
`ExpDisapproval -> Likelihood` effect after controlling for the
*full* set of seven mediators simultaneously. See the speeding
vignette for an extended discussion of why the loop coefficients
(not the per-mediator joint ones) are the right tool for inference
about a single mediator's role, and why `BZ_YX_joint` is best
treated as a system-level diagnostic (Wood, Adanu, & Harms, 2025).

```{r plot-manager-widget, eval = knitr::is_html_output()}
plotPathXMY_widget(dec_m,
                   X_label = "ExpDisapp", Y_label = "Likelihood",
                   Z_label = "manager",
                   Z_levels = c(-0.5, 0, 0.5),
                   format   = "svg")
```

Because `manager` is binary and within-person deviated, the
practical range of `Z` is roughly ±0.5 — toggling `c(-0.5, 0,
+0.5)` approximates "less-manager / average / more-manager"
exposure rather than the SD-based default.

## Situation-level moderation: perceived unethicality (WasUnethical)

A second situation-level moderator comes from `$sit`: `WasUnethical`, rated
on the stem *"At this point in the situation, to what extent has your
[coworker] done something that… Should be regarded as unethical, morally
inappropriate, or 'wrong'?"* (−1 = Greatly Decreased, 0 = Negligible, 1 =
Greatly Increased). Unlike `manager`, which was experimentally assigned,
`WasUnethical` is a continuous participant perception that varies across
situations and persons.

### Data preparation note

Because `$sit` has complete data for only 206 of the 249 participants (43
participants skipped the situation-perceptions block entirely), and because
`WasUnethical` must be merged before within-person deviation so that means
are computed on the actually-available observations, we start from the raw
`$PSI` rather than the pre-deviated `dat`:

```{r prep-unethical}
sit_sub <- coworkerESJT$sit[, c("p","s","WasUnethical")]
dat3 <- merge(coworkerESJT$PSI, sit_sub, by = c("p","s"))
dat3 <- dat3[!is.na(dat3$WasUnethical), ]
cat("Rows:", nrow(dat3), " | Persons:", length(unique(dat3$p)), "\n")

for (v in L1cols) {
  dat3[[v]] <- ave(dat3[[v]], dat3$p, FUN = function(x) x - mean(x))
}
```

Note that within the 206 retained participants, coverage is ~98% (only a
handful of individual situations are missing). The reduction from 5,846 to
4,840 rows is almost entirely due to the 43 persons with no sit data, not
random within-person missingness.

**A caution on standard errors.** `WasUnethical` is a situation-level
perception: both action rows for a given person-situation carry the *same*
value. The cluster-robust SEs produced by `pathXMY()` cluster at the person
level (`p`), which correctly handles person-level non-independence, but do
not additionally account for the extra within-situation correlation introduced
by a shared Z. SEs for paths involving `WasUnethical` may therefore be
modestly optimistic. Two-way clustering (by person and by person × situation)
would address this fully but is not yet implemented.

```{r mod-unethical}
mod_ue <- pathXMY(dat3, X = "ExpDisapproval", Y = "Likelihood",
                  M = mediators, Z = "WasUnethical", Z.within = TRUE)
```

### Does perceived unethicality moderate the expectation arm (ExpDisapproval → M)?

```{r bz-mx-ue}
knitr::kable(
  subset(mod_ue$tidy_loop, param == "BZ_MX")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Moderation of expectation paths by perceived unethicality (BZ_MX)"
)
```

All seven mediators are significantly moderated, and the effects are
substantially larger than those for coworker rank. The dominant pattern is
through `Appropriate` (BZ_MX = 0.599, z = 23.9): in situations perceived as
more unethical than the respondent's personal average, expressing disapproval
is expected to be dramatically *more* role-appropriate. The complementary
picture holds for `Punished` (BZ_MX = −0.307): when the coworker's behavior
is perceived as unethical, the respondent expects *less* personal punishment
for pushing back — the moral justification removes the sense of personal risk.
`CompleteJob` (0.263) and `CompanySuccess` (0.326) also show large positive
moderation: in unethical situations, expressing disapproval is expected to
serve both task-completion and organizational interests far better.

The profile is a near-complete reversal of the manager-rank effect: where a
higher-power coworker suppresses the perceived benefits of asserting,
perceived unethicality amplifies them.

### Does that moderation carry through to the indirect effect?

```{r bz-indirect-ue}
knitr::kable(
  subset(mod_ue$tidy_loop, param == "BZ_MX * B1_YM")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Moderation of indirect effects by perceived unethicality (BZ_MX × B1_YM)"
)
```

The moderated indirect effect through `Appropriate` (0.413) is the largest
such coefficient in this entire vignette — larger than any effect of coworker
rank. In situations perceived as more unethical, the combination of expecting
the action to be appropriate *and* perceived appropriateness strongly driving
likelihood produces a powerful amplification of the indirect path toward
action. `CompanySuccess` (0.218), `Punished` (0.160), and `CompleteJob`
(0.138) also show substantial moderated indirect effects, all reinforcing
the same direction: perceived unethicality makes the full pathway from
expressing-disapproval expectation to action likelihood substantially
stronger.

### Decomposition and field view

```{r decompose-unethical}
dec_ue <- pathXMY_decompose(dat3, X = "ExpDisapproval", Y = "Likelihood",
                            M = mediators, Z = "WasUnethical",
                            Z.within = TRUE)
knitr::kable(dec_ue$total, digits = 3,
             caption = "Total moderation BZ_YX[1] for WasUnethical")
```

```{r plot-unethical, fig.width = 10, fig.height = 7}
plotPathXMY(dec_ue,
            X_label = "ExpDisapp", Y_label = "Likelihood",
            Z_label = "Unethical",
            title = "ExpDisapproval -> M -> Likelihood (Z = WasUnethical)")
```

```{r plot-unethical-widget, eval = knitr::is_html_output()}
plotPathXMY_widget(dec_ue,
                   X_label = "ExpDisapp", Y_label = "Likelihood",
                   Z_label = "Unethical",
                   Z_levels = c(-0.5, 0, 0.5),
                   format   = "svg")
```

Scrub through to watch the field reshape: at low `Unethical` the
`Appropriate` and `CompanySuccess` edges from `ExpDisapproval`
are weak / negative; at high `Unethical` they swing strongly
positive, and the `Likelihood` node shifts from red (don't do it)
toward blue (do it). The whole expectancy-value field flips with
perceived moral justification.

## Person-level moderation: boldness (IIDL_bold)

Does trait boldness moderate these paths? The IIDL item for boldness asks
respondents to rate *"bold, assertive"* on the stem *"How much does each term
describe the way you typically act, feel, or see yourself at work?"*
(−1 = Very Uncharacteristic, 1 = Very Characteristic).

Boldness is a between-person variable, so we use the default `Z.within = FALSE`.
Merging `$IIDL` reduces the sample to the 206 participants who completed the
IIDL measure.

```{r mod-person}
dat2 <- merge(dat, coworkerESJT$IIDL[, c("p","IIDL_bold")], by = "p")
cat("Persons with IIDL_bold:", length(unique(dat2$p)), "\n")

mod_p <- pathXMY(dat2, X = "ExpDisapproval", Y = "Likelihood",
                 M = mediators, Z = "IIDL_bold", Z.within = FALSE)
```

### Does boldness moderate the expectation arm?

```{r bz-mx-bold}
knitr::kable(
  subset(mod_p$tidy_loop, param == "BZ_MX")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Moderation of expectation paths by boldness (BZ_MX)"
)
```

### Does that carry through to the indirect effect?

```{r bz-indirect-bold}
knitr::kable(
  subset(mod_p$tidy_loop, param == "BZ_MX * B1_YM")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Moderation of indirect effects by boldness (BZ_MX × B1_YM)"
)
```

Person-level boldness shows modest moderation overall. The only significant
effect is through `Appropriate` (BZ_MX > 0, p < .05): bolder individuals
expect that directly expressing disapproval will be *more* role-appropriate
than less bold individuals do. This translates into a small but significant
indirect effect: among bolder respondents, expressing disapproval is somewhat
more likely to be taken partly because it is expected to be appropriate.

The contrast with the situation-level results is informative: *where* you are
in the power hierarchy (as captured by the randomly assigned coworker rank)
shapes expected outcomes far more pervasively than *who* you are in terms of
self-reported boldness.

### Decomposition and field view

```{r decompose-bold}
dec_b <- pathXMY_decompose(dat2, X = "ExpDisapproval", Y = "Likelihood",
                           M = mediators, Z = "IIDL_bold",
                           Z.within = FALSE)
knitr::kable(dec_b$total, digits = 3,
             caption = "Total moderation BZ_YX[1] for IIDL_bold")
```

```{r plot-bold, fig.width = 7, fig.height = 3.5}
plotPathXMY(dec_b, mediator = "Appropriate",
            X_label = "ExpDisapp", Y_label = "Likelihood",
            Z_label = "bold",
            title = "ExpDisapproval -> Appropriate -> Likelihood (Z = bold)")
```

```{r plot-bold-widget, eval = knitr::is_html_output()}
plotPathXMY_widget(dec_b, mediator = "Appropriate",
                   X_label = "ExpDisapp", Y_label = "Likelihood",
                   Z_label = "bold",
                   format  = "svg")
```

The widget toggles `bold = -1, 0, +1` (one SD below the mean,
mean, one SD above, on the z-standardized between-person scale).
The change is subtle compared to the situation-level moderators —
bolder respondents see expressing disapproval as more appropriate,
but every other path is essentially fixed.

## Function reference

See `?pathXMY`, `?pathXMY_decompose`, `?plotPathXMY`,
`?plotPathXMY_widget`, `?esjtCov`, and `?coworkerESJT`.
