---
title: "Notation: F, F_t, and the fZ_X.Y schema"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Notation: F, F_t, and the fZ_X.Y schema}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

This vignette pins down the notation used by every other piece of the
package: how a functional field is written, how its parameters are
labelled, and how those labels map onto `lavaan` model strings. It is
short on prose and long on conventions — the goal is a fixed target
that other vignettes and helpers can refer back to without re-deriving.

## Two objects with the same letter

A **functional field** appears in two forms:

- **`F`** — the *time-invariant* field. A square matrix of equations,
  indexed `F[X, Y]` = the force *from* source `X` *to* target `Y`. Each
  cell holds an expression in the situational state `s`; in practice
  those expressions are simple linear combinations of state variables,
  but in principle they can be arbitrary.

- **`F_t = F(s_t)`** — the *time-varying* field. The numerical 2-matrix
  obtained by evaluating each cell of `F` at the current state `s_t`.

The Markov step has two sub-steps, both keyed on `s_t`:

\[
F_t   = F(s_t) \qquad
s_{t+1} = s_t \cdot F_t.
\]

`F` is fixed; `F_t` changes from step to step because `s_t` does.
Actions enter the loop by perturbing `s_t` (`s_t[i] \leftarrow s_t[i] + \delta`).
Iteration continues until the field stops producing change, at which
point a designated **selection signal** `L_s` (typically rated
likelihood in ESJT data) is read off.

The source-then-target indexing (`F[X, Y]`) is deliberate and
contrasts with the standard SEM `B[Y, X]` convention. Anywhere a
coefficient is extracted from `lavaan` and exposed as an `F`-cell, it
is transposed at the boundary.

## The parameter schema

Every estimable parameter in `F` carries a label of the form

\[
\boxed{\texttt{fZ\_X.Y}}
\]

Read as: the **f**-convention coefficient on the force from `X` to
`Y`, under moderator `Z`. The schema is universal — every parameter
fits this template — and is regular in two ways:

1. The **baseline** (unmoderated) coefficient is the special case
   `Z = 1`, written `f1_X.Y`. It is the "interaction of `X` with the
   constant 1" — i.e., the part of the force that fires regardless of
   the situation. We keep the `1` explicit; there is no bare
   `f_X.Y`.
2. Multiple moderators on the same cell are written as multiple
   `fZ_X.Y` terms with different `Z`. There is no built-in limit on
   how many a cell can carry.

The separators are load-bearing: `_` separates the moderator from the
path, and `.` separates source from target within the path. Both `_`
and `.` are valid R identifier characters, so `lavaan` parses the
labels intact (verified against lavaan 0.6-21).

**Generic single-letter slots drop the separator.** When the source and
target are the one-letter schematic roles `X`, `M`, `Y` — as they always
are on the *inductive* side, where `pathXMY()` fits a generic
`X → M → Y` template — the `.` is unnecessary and the keys are written
without it: `f1_XM`, `fZ_MY`, `fZ_XY` (read the two trailing letters as
source then target). The dotted form is reserved for *named* nodes, as
on the *deductive* side (`fS2_TurnOn.HCoPot`), where multi-character
names would otherwise run together. Both forms render to the same
matrix-cell display, `F[src,tgt]` — `fZ_XM` and `FZ[X,M]` are the same
parameter, one a lavaan-safe key, the other a reading aid (see
[The F1 / FZ matrix overlay](#the-f1-fz-matrix-overlay) and
`pathXMY_to_F()`).

### Reading a cell

A cell `F[X, Y]` with a baseline term and one Z-moderation expands as

\[
F[X, Y] = f1\_X.Y \;+\; fZ\_X.Y \cdot s[Z].
\]

Evaluated at `s_t`:

\[
F_t[X, Y] = f1\_X.Y \;+\; fZ\_X.Y \cdot s_t[Z].
\]

Read into the propagation step: the contribution of source `X` to
target `Y` at time `t` is `s_t[X] · F_t[X, Y]`. So `Y`'s update,
summing over all sources, is

\[
s_{t+1}[Y] \;=\; \sum_X s_t[X] \cdot F_t[X, Y].
\]

## Lavaan-syntax mapping

A cell `F[X, Y]` with a baseline term and a `Z`-moderation term
becomes one row in a `lavaan` model:

```
Y ~ f1_X.Y * X + fZ_X.Y * Z:X
```

The target sits on the left of `~` (lavaan's regression convention);
the source sits on the right; the moderator enters as the leading
member of a `Z:X` interaction. The interaction is written **condition
first, action last** so its order matches the label: `Z:X` ↔ `fZ_X.Y`.
This is load-bearing for `labelF()`, which auto-generates the labels by
reading the final right-hand variable as the source/action and any
leading ones as the condition. The parameter labels record the
F-convention semantics regardless of `lavaan`'s left/right orientation.

A small worked check — the model parses cleanly and the labels
survive `lavaanify()`:

```{r lavaanify-check}
library(lavaan)
mod <- "Y ~ f1_X.Y * X + fZ_X.Y * Z:X"
pt  <- lavaanify(mod, fixed.x = FALSE)
pt[pt$op == "~", c("lhs", "op", "rhs", "label")]
```

A cell with *only* a Z-moderation term and no baseline is an
**afforded force**: the source `X` contributes nothing to `Y` when
`s[Z] = 0`, and the force becomes available once `s[Z]` flips on. This
is how the redesigned coffee vignette expresses
prerequisites-must-hold actions:

```
HCoPot ~ fS2_TurnOn.HCoPot * s2:TurnOn
```

The `TurnOn → HCoPot` force is zero until `s2` is flipped on by an
earlier action. The Markov iteration handles the rest.

## Situation, action plan, and fixed coefficients

The forces above describe the **situation** — the affordance structure
of the world, independent of any agent. Layered on top is an **action
plan**: conditional commitments of the form "fire action `A` once state
`Z` is achieved", written as state → action forces:

```
make   ~ 1 * s1
TurnOn ~ 1 * s2
```

Two notational points. First, plan forces run *opposite* to situation
forces — state → action, where the situation runs action → outcome.
Second, the coefficient is **fixed** (`1 * s1`, lavaan's
fixed-coefficient syntax) rather than carrying an `fZ_X.Y` label: a plan
trigger is a structural commitment, not a quantity estimated from data,
so it lives outside the schema and needs no `params` entry. `evalF()`
values a fixed term straight from its coefficient.

Because both halves are plain `lavaan` strings, the field is assembled
by concatenation — `paste(situation, plan)` — and run as one model. By
visualization convention the plan's forces are drawn in **gold**
(`plotField(..., plan = plan)`), marking the forces a conditional action
plan adds on top of the situation. Splitting the field this way takes
about twice as many iterations to settle (each stage is now two hops,
state → action → outcome), but reaches the same fixed point as injecting
the actions on an external schedule.

## The F1 / FZ matrix overlay

A field's parameters can be reorganized into a stack of per-moderator
matrices:

- `F1`: cell `(i, j) = f1_i.j` — the baseline force matrix.
- `FZ`: cell `(i, j) = fZ_i.j` — the contribution to `F` that is
  proportional to `s[Z]`.
- `FZ'`, `FZ''`, ... — one matrix per additional moderator.

Each matrix is itself an indexable view onto the underlying parameter
table. Visualization overlays them — for example, render `F1` as the
black edges that are always active, and `FZ` as colored edges that
strengthen with `s[Z]`. The three-panel decomposition figures in the
speeding `pathXMY` vignette are an early hand-built instance of this;
generalising the data side via `pathXMY_to_F()` lets the same
visualization machinery render whole fields rather than single
triangles.

The matrix view is also the convenient handoff to the runtime: a step
`s_{t+1} = s_t \cdot F_t` decomposes as

\[
F_t \;=\; F_1 \;+\; \sum_Z s_t[Z] \cdot F_Z,
\]

so `F_t` is a state-weighted sum of the overlay matrices. This is
exactly the form `evalF()` consumes.

## What the schema does *not* impose

The label convention is purely a naming discipline. It does not by
itself commit any cell to a particular functional form:

- A cell can be a **constant** (`f1_X.Y` only) — a topological edge,
  fit by an ordinary path coefficient.
- A cell can be **state-linear** (`f1_X.Y` + one or more `fZ_X.Y`
  terms) — a moderated edge, fit by regression with interactions.
- A cell can be an **afforded force** (one `fZ_X.Y` term, no
  `f1_X.Y`) — fires only when `s[Z] > 0`. The structural-masking
  trick the coffee vignette relies on.
- A cell can carry **multiple moderators** simultaneously by stacking
  `fZ_X.Y` terms; the linear-combination interpretation extends.

Nonlinearities like `s[a] \cdot s[b]` gates appear as `a:b:X`
interaction terms in the lavaan-syntax mapping (conditions first,
action last), with a label of the form `fab_X.Y` (compound moderator).
The package treats compound moderators as one logical "Z" — the
matrix-overlay view holds an extra `F_{ab}` matrix; nothing else
changes.

## `f1_X.Y` as a placeholder

Read strictly, every cell of `F` is afforded: there are no truly
unconditional forces. When `f1_X.Y` appears in a model, it captures
the X → Y force *under whatever conditions held during the relevant
data or theory* — without naming those conditions. The "1" is a
shorthand for "a state variable that happens to be true throughout
the data we have." So a `f1_X.Y` term is best read as a frank
acknowledgement that a force is present in these conditions, and
that the conditions that gate it have not yet been identified.

A mature representation of a process replaces every `f1_X.Y` with
one or more `fZ_X.Y` for the `Z`'s that determine when the force
fires — and which, in principle, can be intervened on to interrupt
it. The schema admits `f1_X.Y` as a convenience (forcing full
conditional specification at every iteration would stall any working
model), but the conceptual direction of progress is to push every
`f1` toward an `fZ` decomposition over time. A discovered "main
effect" is, on this reading, a moderation effect waiting for its
moderator to be named.

## Node-name conventions

Names that appear on either side of `~`, or inside an `X:Z`
interaction, must be lavaan-safe: letters, digits, and `_` only. No
parens, no `.`, no other punctuation. The `.` is reserved as the
source/target separator inside parameter labels.

This is a change from the original `coffee_field_model` vignette,
which used names like `H(Co).Pot` and `make(s2).p`. The redesigned
vignette renames these to `HCoPot` and `make_s2`. The pretty
display labels (with parens and dots) remain available for
visualization — `prepField()` and the plotting helpers carry a
separate display-label map.

## The two roles a field can play

The same `F` lives at the centre of two complementary workflows:

| Role | Direction | Tool | Parameters |
|---|---|---|---|
| **Induction** — learn the world from observed behaviour | Data → `F` | `pathXMY()` and friends | Fit, uncertain, carry SEs |
| **Deduction** — use a working theory of the world to plan | `F` → behaviour | `simulateF()`, `chooseF()`, the coffee vignette | Hand-set, treated as known |

Both directions use the same model syntax and the same parameter
schema. The bridge between them — turning a `pathXMY` fit into a
runnable field — is `pathXMY_to_F()`.

## Migration from the legacy `B1_MX` / `BZ_MX` labels

Earlier versions of `pathXMY()` labelled their `tidy_loop` /
`tidy_joint` tables in the SEM-style `B1_MX` / `BZ_YX` convention
(prefix `B`, scope digit, *target-source* suffix). The tables now key on
the F-schema generic-slot form directly — prefix `f`, scope digit,
*source-target* suffix, no separator (single-letter slots). The bracket
column is the display form produced by `pathXMY_to_F()`:

| Legacy label | F-schema key | Display | Reads as |
|---|---|---|---|
| `B1_MX`  | `f1_XM` | `F1[X,M]` | baseline X → M force |
| `BZ_MX`  | `fZ_XM` | `FZ[X,M]` | Z-moderation of X → M |
| `B1_YM`  | `f1_MY` | `F1[M,Y]` | baseline M → Y force |
| `BZ_YM`  | `fZ_MY` | `FZ[M,Y]` | Z-moderation of M → Y |
| `B1_YX`  | `f1_XY` | `F1[X,Y]` | direct X → Y (controlling for M) |
| `BZ_YX`  | `fZ_XY` | `FZ[X,Y]` | Z-moderation of direct X → Y |

The joint multi-mediator rows carry a `_joint` suffix (`f1_XY_joint`,
`fZ_XY_joint`, …); the no-mediator *total* is written `f1_XY` with a
`*` superscript in display, `F1*[X,Y]`. `pathXMY_to_F()` (see its own
help page) renders any of these keys — singles, products
(`fZ_XM * f1_MY` → `FZ[X,M]·F1[M,Y]`), and `_joint` / `_total` /
`(direct)` variants — into the bracket display used throughout the
`pathXMY` vignettes. The deductive engine keeps the dotted, named-node
form `fZ_X.Y` (above); the two coincide once nodes are single-letter
slots.
