---
title: "Moderated mediation with `pathXMY()`: the work-overtime EXSJT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Moderated mediation with pathXMY(): the work-overtime EXSJT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png",
  dpi = 192,
  out.width = "70%"
)
options(width = 120)
```

This vignette walks through moderated mediation with `pathXMY()` using
the `overtimeESJT` dataset — an **EXSJT** in which a single
work-overtime scenario was randomly varied across eight features, and
each participant rated two actions (agree to work overtime; decline) on
nine expected outcomes.

**Same respondents as `speedingESJT`.** The work-overtime, speeding, and
data-home EXSJTs were all administered to the same Prolific panel
(N = 347). The three datasets (`overtimeESJT`, `speedingESJT`,
`dataHomeESJT`) share a participant pool and can be linked by `p` —
each respondent's id appears in all three datasets that retained them
after per-block quality screens. This makes cross-scenario consistency
questions (e.g., do trait-level differences predict similar
expectation shifts across domains?) tractable without recollecting
data.

## The data

```{r data}
library(funfield)
data(overtimeESJT)
lapply(overtimeESJT, function(x) if (is.data.frame(x)) dim(x) else length(x))
```

## Variables and survey wording

### The scenario

Each participant saw exactly one scenario, randomly assembled from eight
situational factors. The Qualtrics scenario template lives at
`overtimeESJT$codebook$scenario$template`; what that returns is shown
below in a readable format. The `${...}` placeholders are filled in
per respondent from the eight factors that follow:

```{r scenario-template, echo = FALSE, results = "asis"}
cat("> ", overtimeESJT$codebook$scenario$template, "\n")
```

```{r scenario-factors}
knitr::kable(
  overtimeESJT$codebook$scenario$factors,
  col.names = c("Factor", "Levels (semicolon-separated)"),
  caption = "The eight factors that varied across respondents"
)
```

Each respondent's fully assembled scenario is in
`overtimeESJT$cond$scenario`. Below are the first three (i.e., what
`overtimeESJT$cond$scenario[1:3]` returns, again formatted for
readability):

```{r scenario-assembled, echo = FALSE, results = "asis"}
ss <- overtimeESJT$cond$scenario[1:3]
for (i in seq_along(ss)) {
  cat("> **Respondent ", i, ".** ", ss[i], "\n\n", sep = "")
}
```

Notice how the same template produces materially different scenarios
through random factor assignment. The factor levels that were combined
to produce each of the three scenarios above:

```{r scenario-cond}
knitr::kable(
  overtimeESJT$cond[1:3, c("CoType","weekday","priorPlans","workDifficulty",
                           "cashReward","projectImpact")],
  caption = "Factor levels for respondents 1-3"
)
```

### Expected outcomes

After reading the scenario, each respondent rated each of the two
candidate actions on the nine expected-outcome items below
(0 = Very Low … 1 = Very High):

```{r wording-table}
knitr::kable(
  overtimeESJT$codebook$outcomes[, c("var", "label")],
  col.names = c("Variable", "Survey wording"),
  caption = "Outcome items, rated on [0, 1]"
)
```

The focal **predictor** is `Overtime` — the action coding
(0 = decline, 1 = agree to work overtime). The focal **outcome** is
`Likelihood`. The nine **mediators** are the expected-outcome ratings.

## Data preparation

Within-person deviate the Level-1 columns.

```{r prep}
L1cols <- c("Overtime","RelationshipQuality","EmployerValuation",
            "OutsideRelationships","Drained","CompanyHealth","Finances",
            "Appropriate","RaisePromotion","Punished","Likelihood")
mediators <- c("RelationshipQuality","EmployerValuation","OutsideRelationships",
               "Drained","CompanyHealth","Finances","Appropriate",
               "RaisePromotion","Punished")

dat <- overtimeESJT$PSI
for (v in L1cols) {
  dat[[v]] <- ave(dat[[v]], dat$p, FUN = function(x) x - mean(x))
}
```

## Normative model (no moderator)

```{r normative}
norm <- pathXMY(dat, X = "Overtime", Y = "Likelihood", M = mediators)
```

### Expectation paths (F1[X,M]): what does agreeing to overtime produce?

```{r bmx}
knitr::kable(
  subset(norm$tidy_loop, param == "f1_XM")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Expectation paths (Overtime -> M)"
)
```

Agreeing to work overtime is expected to substantially increase
`Drained` (+0.32), `RelationshipQuality` with the asker (+0.28),
`RaisePromotion` (+0.21), and `EmployerValuation` (+0.16). It is
expected to reduce `OutsideRelationships` (−0.16) and the
likelihood of being `Punished` by the company (−0.16). Smaller
positive expectations attach to `CompanyHealth`, `Finances`, and
`Appropriate`.

### Valuation paths (F1[M,Y]): which expectations drive action likelihood?

```{r bym}
knitr::kable(
  subset(norm$tidy_loop, param == "f1_MY")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Valuation paths (M -> Likelihood)"
)
```

`Appropriate` (+0.90) and `Finances` (+0.89) are the strongest
positive drivers of taking the action. `RaisePromotion` (+0.65),
`EmployerValuation` (+0.60), and `RelationshipQuality` (+0.57) also
strongly increase likelihood. `Drained` (−0.33) and `Punished`
(−0.35) suppress likelihood. The valuation pattern is intuitive:
financial, evaluative, and relational gains pull respondents toward
the action; expected fatigue and punishment push them away.

### Indirect effects (F1[X,M]·F1[M,Y])

```{r bind}
knitr::kable(
  subset(norm$tidy_loop, param == "f1_XM * f1_MY")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Indirect effects through each mediator"
)
```

The pull toward agreeing comes mainly through expected gains in
`RelationshipQuality` (+0.16), `RaisePromotion` (+0.14), `Finances`
(+0.11), and `EmployerValuation` (+0.10). The push toward declining
comes through expected `Drained` (−0.10) and harm to
`OutsideRelationships` (−0.06).

## Situation-level moderation: is overtime paid?

`WO.cashReward` was randomly assigned at the person level: either the
company pays for the overtime work, or it doesn't. Code as a binary
`paid` indicator and merge:

```{r prep-paid}
cot <- overtimeESJT$cond[, c("p","cashReward")]
cot$paid <- as.numeric(cot$cashReward ==
  "the company will pay you overtime (above your normal pay rate) to work on the project")
dat <- merge(dat, cot[, c("p","paid")], by = "p")

mod_s <- pathXMY(dat, X = "Overtime", Y = "Likelihood",
                 M = mediators, Z = "paid", Z.within = FALSE)
dec_paid <- pathXMY_decompose(dat, X = "Overtime", Y = "Likelihood",
                              M = mediators, Z = "paid", Z.within = FALSE)
```

### Can we identify reasons that paying for overtime increases the likelihood of agreeing to work overtime?

Pay status's effect on the overall agreement decision routes
through *which* expected outcomes get shifted (the **expectation
arm**, `FZ[X,M]`) and through *how* each expected outcome is
weighted on the choice (the **valuation arm**, `FZ[M,Y]`). The
pretty pairtable lays the two arms side by side:

```{r bz-pair-paid}
group_kable(
  pathXMY_pairtable(mod_s, c("fZ_XM", "fZ_MY")),
  groups = c(" " = 1,
             "Pay-status moderation: expectation arm (FZ[X,M])" = 4,
             "Pay-status moderation: valuation arm (FZ[M,Y])"   = 4),
  col_labels = c("Mediator", "est", "se", "z", "p",
                             "est", "se", "z", "p")
)
```

On the **expectation arm**, six of the nine outcomes are
significantly moderated. The largest shifts are in `Appropriate`
(+0.141, *p* < .001) and `Finances` (+0.121, *p* < .001) —
agreeing to paid overtime feels both more role-appropriate and
genuinely produces more money, by a large margin. Smaller but
significant positive shifts attach to `EmployerValuation`,
`OutsideRelationships`, `RelationshipQuality`, and
`RaisePromotion`. The cost-side outcomes — `Drained`,
`Punished`, `CompanyHealth` — are untouched: pay status doesn't
change how depleting or risky people expect overtime to be, it
shifts the *upside* of the action.

On the **valuation arm**, only one path reaches *p* < .05:
`Drained` (+0.190, *p* = .02). The positive sign means the
average negative weight that fatigue carries on the
agreement decision (`F1[M,Y]` ≈ −0.33) is *less* negative when
overtime is paid — being tired matters less to the choice when
there's money on the table. Two cost paths near-miss
significance in the same direction (`Punished`, *p* = .08;
`OutsideRelationships`, *p* = .08). As in the speeding fits,
the valuation-arm standard errors run roughly 5× the
expectation-arm SEs (a design-variance asymmetry: `X` has
guaranteed within-person variation, the mediators don't always
follow), so magnitude-equivalent effects on the right side are
held to a weaker statistical claim than on the left.

#### Z-overlay field

The same coefficients as a **Z-overlay field** — the normative
layout, but with every `F1` swapped for its `FZ` counterpart.
Each edge is the per-unit-`paid` change in that path's weight;
near-zero coefficients fade to light gray. Because `paid` is
binary and z-standardized inside `pathXMY()`, each edge is half
the no-pay → pay swing in the corresponding F1.

```{r plot-overlay-paid, fig.width = 9, fig.height = 6}
plotPathXMY(mod_s,
            X_label = "Overtime", Y_label = "Likelihood",
            Z_label = "paid", X_shape = "rtTri",
            Z_overlay = TRUE,
            scale_max = 0.3,
            title = "Z overlay: how pay status shifts path weights")
```

Visually the asymmetry is obvious: the strongest moderation
sits on the left half of the diagram (expectation arms into
`Appropriate` and `Finances`), with most of the right half
(valuation arms out of the mediators) faded out.

#### Do those shifts create reasons to agree?

The expectation arm and the valuation arm only become *reasons*
when paired with their counterpart on the other side. An
expectancy shift through `Appropriate` is only a reason if
`Appropriate` carries weight on `Likelihood`; a valuation shift
on `Drained` is only a reason if overtime actually moves
`Drained`. `plotPathXMY_routes()` shows the two products
directly: the **expectation route** pairs each green
`FZ[X,M](paid)` shift with the surviving black average
`F1[M,Y]`; the **valuation route** pairs each black average
`F1[X,M]` with each green `FZ[M,Y](paid)`. Both panels are read
left-to-right: any complete chain where both arms are visibly
thick is a candidate reason that paying shifts agreement.

```{r plot-routes-paid, fig.width = 14, fig.height = 7}
plotPathXMY_routes(dec_paid,
                   X_label = "Overtime", Y_label = "Likelihood",
                   Z_label = "paid", X_shape = "rtTri",
                   scale_max = 0.5)
```

The **expectation route** does most of the work. The strongest
reason runs through `Appropriate` (+0.099, *p* < .001) —
agreeing to *paid* overtime feels appropriate to a degree that
agreeing to unpaid overtime does not, and appropriateness in
turn strongly drives the action. The second-largest reason is
`Finances` (+0.054, *p* < .001): paid overtime is expected to
actually produce financial gain (it's the definition of the
condition), and financial gain weighs heavily on the choice.
Smaller but real expectation-route reasons run through
`EmployerValuation` (+0.022, *p* = .01) and `RaisePromotion`
(+0.016, *p* = .02), with `RelationshipQuality` and
`OutsideRelationships` borderline.

The **valuation route** contributes one significant reason:
`Drained` (+0.060, *p* = .02). Overtime is reliably expected
to produce fatigue (`F1[X,M](Drained)` ≈ +0.32 normatively); when
that fatigue's *weight* on the decision shrinks under paid
conditions, that shrinkage acts as a route toward agreement
even though the underlying expectation didn't change.
`Punished` and `OutsideRelationships` carry borderline
valuation-route reasons in the same direction (when paid, the
threat of punishment and the threat to outside relationships
matter slightly less to the choice).

The composite picture: paying for overtime increases agreement
mostly by changing what respondents *expect* the action to
produce — primarily that it is appropriate to agree, and that
agreement will actually be financially rewarding — and
secondarily by softening the *weight* placed on the one cost
the action reliably produces (fatigue). The upsides get
amplified and one of the downsides gets discounted; the rest
of the cost ledger stays as it was.

### Decomposition and field view

```{r decompose-paid}
knitr::kable(dec_paid$total, digits = 3,
             caption = "Total moderation FZ*[X,Y] for pay status")
```

```{r plot-paid, fig.width = 10, fig.height = 7}
plotPathXMY(dec_paid,
            X_label = "Overtime", Y_label = "Likelihood",
            Z_label = "paid",
            X_shape = "rtTri",
            title = "Overtime -> M -> Likelihood (Z = paid)")
```

The fan diagram's mediator arms are the per-mediator *loop* fits
(one X to M to Y model per mediator, overlaid). The straight
**X to Likelihood arrow** is the residual direct path from the
*joint* multi-mediator fit (added automatically by
`pathXMY(joint = TRUE)`, the default) — i.e. the
`Overtime -> Likelihood` effect after controlling for the *full*
set of nine mediators simultaneously. See the speeding vignette
for an extended discussion of why the loop coefficients (not the
per-mediator joint ones) are the right tool for inference about
a single mediator's role, and why the joint `FZ[X,Y]` is best treated
as a system-level diagnostic (Wood, Adanu, & Harms, 2025).

```{r plot-paid-widget, eval = knitr::is_html_output()}
plotPathXMY_widget(dec_paid,
                   X_label = "Overtime", Y_label = "Likelihood",
                   Z_label = "paid",
                   X_shape = "rtTri",
                   Z_levels = c(-1, 1),
                   format   = "svg")
```

`paid` is binary (no / yes) and z-standardized inside `pathXMY()`,
so `Z = -1` and `Z = +1` map exactly to "no pay offered" and "pay
offered." The widget uses just those two frames. Toggling between
them: the upper arms (`Finances`, `Appropriate`,
`EmployerValuation`, `RaisePromotion`) all thicken when paid, while
the cost edges (`Drained`, `Punished`) stay fixed.

## Person-level moderation: vTime

`vTime` is the single-item self-report *"I value my time"* (SR_12 in
`$traits`, rescaled to [0, 1]). As a between-person trait, it uses
`Z.within = FALSE`.

```{r prep-vtime}
tr <- overtimeESJT$traits[, c("p","SR_12")]
names(tr)[2] <- "vTime"
tr <- tr[!is.na(tr$vTime), ]
dat2 <- merge(dat, tr, by = "p")

mod_p <- pathXMY(dat2, X = "Overtime", Y = "Likelihood",
                 M = mediators, Z = "vTime", Z.within = FALSE)
```

### Does vTime moderate the expectation arm?

```{r bzmx-vtime}
knitr::kable(
  subset(mod_p$tidy_loop, param == "fZ_XM")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Moderation of expectation paths by vTime (FZ[X,M])"
)
```

Respondents who report valuing their time more expect agreeing to
overtime to produce *larger* gains in `EmployerValuation`,
`CompanyHealth`, `Finances`, and `RaisePromotion`. Each effect is
modest but the pattern is coherent: high-vTime respondents view
overtime as a more profitable investment of their time, both
financially and reputationally.

### Does that carry through to the indirect effect?

```{r bz-indirect-vtime}
knitr::kable(
  subset(mod_p$tidy_loop, param == "fZ_XM * f1_MY")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Moderation of indirect effects by vTime (FZ[X,M]·F1[M,Y])"
)
```

The corresponding indirect-path moderation is in the same direction
for the same mediators, though only the largest ones reach
significance after the multiplication by the valuation weights.

### Decomposition and field view

```{r decompose-vtime}
dec_vt <- pathXMY_decompose(dat2, X = "Overtime", Y = "Likelihood",
                            M = mediators, Z = "vTime", Z.within = FALSE)
knitr::kable(dec_vt$total, digits = 3,
             caption = "Total moderation FZ*[X,Y] for vTime")
```

```{r plot-vtime, fig.width = 7, fig.height = 3.5}
plotPathXMY(dec_vt, mediator = "Finances",
            X_label = "Overtime", Y_label = "Likelihood",
            Z_label = "vTime",
            X_shape = "rtTri",
            title = "Overtime -> Finances -> Likelihood (Z = vTime)")
```

```{r plot-vtime-widget, eval = knitr::is_html_output()}
plotPathXMY_widget(dec_vt, mediator = "Finances",
                   X_label = "Overtime", Y_label = "Likelihood",
                   Z_label = "vTime",
                   X_shape = "rtTri",
                   format  = "svg")
```

The vTime moderation is subtle but consistent: as time-valuing
rises, the `Overtime -> Finances` edge slightly thickens (people
who value time more expect a larger financial return from
agreeing). The other paths in the triangle are essentially fixed.

## Quickly screening for moderator candidates

`vTime` was a single guess about which trait might shift the
overtime decision. The traits table holds 30 self-report items
plus age, and the situation block adds six experimentally varied
factors — most untouched so far. Fitting a separate
`pathXMY()` for each one would be expensive, but a shortcut
falls out of the structure of the total moderation itself: the
no-mediator `FZ*[X,Y]` from a fit with a between-person
candidate `Z` is algebraically equivalent (up to an SD-of-slope
constant that cancels) to the Pearson correlation between `Z`
and each person's within-person `Overtime → Likelihood` slope.
`screenModerators()` packages this shortcut and reports the
estimate on the `FZ` scale; see the speeding vignette for the
derivation and the discussion of how its *p*-value relates to
the cluster-robust *p* from `pathXMY()`.

### Trait screen

```{r screen-traits-overtime}
screen <- screenModerators(psi    = dat,
                           X      = "Overtime",
                           Y      = "Likelihood",
                           M      = mediators,
                           traits = overtimeESJT$traits)

labels <- overtimeESJT$codebook$trait_items[, c("var","label")]
screen <- merge(screen, labels,
                by.x = "trait", by.y = "var", all.x = TRUE, sort = FALSE)

screen_yx <- subset(screen, target == "YX")
screen_yx <- screen_yx[order(-abs(screen_yx$beta)), ]
knitr::kable(head(screen_yx, 15)[, c("trait","r","beta","p","label")],
             digits = 3, row.names = FALSE,
             caption = paste0("Top 15 traits ranked by the within-person ",
                              "Overtime effect on Likelihood: a fast screen ",
                              "for total-moderation candidates."))
```

The top of the trait list is unusually coherent. The strongest
single-item moderator is *"I am compassionate, have a soft
heart"* (`SR_18`, *r* = .24, *p* &lt; .001), followed by *"I try
to follow the rules"* (`SR_01`), *"I assume the best about
people"* (`SR_20`), *"I would be upset if I were unemployed"*
(`SR_10`), age, *"I am fascinated by art, music, or literature"*
(`SR_27`), *"I am reliable, can always be counted on"*
(`SR_23`), *"I am helpful and unselfish with others"* (`SR_14`),
and *"I take commitments I have made to people seriously"*
(`SR_09`). The lone item with a negative sign in the top fifteen
is *"I am sometimes rude to others"* (`SR_19`) — the opposite
end of the same agreeable continuum.

Read together, this is a pro-social / conscientious /
agreeable cluster: respondents who agree to work overtime in
this scenario are not, item by item, distinguished by valuing
their time or by ambition for the raise — they are distinguished
by self-descriptions oriented toward duty, dependability, and
care for the people who would be affected. `vTime` (`SR_12`)
itself sits well outside the top fifteen (*r* = −.06, *p* = .31),
which makes the choice of `vTime` as a moderator above a
genuinely cold-pick illustration rather than an *a priori*
favorite.

### Situational factor screen

The six randomly varied situational factors slot into the same
`screenModerators()` call once they are recoded to 0/1
indicators — the `traits` argument accepts any per-person frame:

```{r screen-situation-overtime}
cond <- overtimeESJT$cond
sit_num <- data.frame(
  p              = cond$p,
  paid           = as.numeric(cond$cashReward ==
                              "the company will pay you overtime (above your normal pay rate) to work on the project"),
  boss           = as.numeric(cond$CoType == "Your boss"),
  friday         = as.numeric(cond$weekday == "Friday evening"),
  priorPlans     = as.numeric(cond$priorPlans ==
                              "have previously made plans to meet with friends"),
  difficult      = as.numeric(cond$workDifficulty == "difficult"),
  helpfulProject = as.numeric(cond$projectImpact ==
                              "is clear that completing this project will be very helpful to the company")
)

screen_sit <- screenModerators(psi = dat, X = "Overtime", Y = "Likelihood",
                                M = mediators, traits = sit_num)
screen_sit_yx <- subset(screen_sit, target == "YX")
screen_sit_yx <- screen_sit_yx[order(-abs(screen_sit_yx$beta)), ]
knitr::kable(screen_sit_yx[, c("trait","r","beta","p","n")],
             digits = 3, row.names = FALSE,
             caption = "Situational factors ranked by their moderation of Overtime -> Likelihood")
```

Pay status dominates the situational list — and the combined
list. Toggling from unpaid to paid overtime shifts the
within-person `Overtime → Likelihood` slope by **+.34**, more
than twice the strongest trait moderator. Having previously made
plans with friends pulls the slope down by .15 (*p* &lt; .001),
and a "difficult" framing of the work pulls it down by a further
.08 (*p* = .03). The remaining three situational factors —
whether it is the boss or a coworker doing the asking, whether
it is a Friday or Wednesday evening, and whether the project's
helpfulness to the company is clear — do not reach significance
on their own.

### Largest FZ[X,Y] moderators, combined ranking

Putting the two screens together names the variables that should
carry the largest `FZ[X,Y]` coefficients in a full
`pathXMY_decompose()` fit on this dataset:

| Rank | Variable                                          | Type        |    β    |
|-----:|---------------------------------------------------|-------------|--------:|
| 1    | overtime is paid                                  | situation   |  +.343  |
| 2    | *"I am compassionate, have a soft heart"*         | trait       |  +.146  |
| 3    | prior plans with friends                          | situation   |  −.147  |
| 4    | *"I try to follow the rules"*                     | trait       |  +.130  |
| 5    | *"I assume the best about people"*                | trait       |  +.117  |
| 6    | *"I would be upset if I were unemployed"*         | trait       |  +.080  |
| 7    | work described as difficult                       | situation   |  −.076  |
| 8    | age                                               | demographic |  +.075  |

The dominant single moderator is the experimentally manipulated
pay indicator — a satisfying result, in that the scenario was
*designed* to surface it. The trait moderators that fall out of
the screen aren't the candidates a researcher might guess from
the overtime-as-investment framing (`vTime`, ambition, ratings
of one's leadership) but a tight cluster of agreeable /
conscientious / pro-social self-descriptions. That cluster is a
natural next target for a per-trait `pathXMY_decompose()` —
each item rephrases the same underlying disposition, so the
field decompositions should look similar across them and the
exercise becomes one of identifying which mediators carry that
shared moderation rather than which item to model first.

## Cross-scenario linkage

`overtimeESJT$traits` and `speedingESJT$traits` carry the same 30
self-report items for the same set of respondents. Cross-domain
analyses can be done by joining on `p`:

```{r cross-scenario, eval = FALSE}
data(speedingESJT)
both <- merge(
  overtimeESJT$traits[, c("p","SR_12")],
  speedingESJT$traits[, c("p","SR_12")],
  by = "p"
)  # identical SR_12 within p, by design — confirms the same-respondents claim
```

## Function reference

See `?pathXMY`, `?pathXMY_decompose`, `?plotPathXMY`,
`?plotPathXMY_widget`, `?esjtFromQualtrics`, and `?overtimeESJT`.
