---
title: "Moderated mediation with `pathXMY()`: the work-overtime EXSJT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Moderated mediation with pathXMY(): the work-overtime EXSJT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png",
  dpi = 192,
  out.width = "70%"
)
options(width = 120)
```

This vignette walks through moderated mediation with `pathXMY()` using
the `overtimeESJT` dataset — an **EXSJT** in which a single
work-overtime scenario was randomly varied across eight features, and
each participant rated two actions (agree to work overtime; decline) on
nine expected outcomes.

**Same respondents as `speedingESJT`.** The work-overtime, speeding, and
data-home EXSJTs were all administered to the same Prolific panel
(N = 347). The three datasets (`overtimeESJT`, `speedingESJT`,
`dataHomeESJT`) share a participant pool and can be linked by `p` —
each respondent's id appears in all three datasets that retained them
after per-block quality screens. This makes cross-scenario consistency
questions (e.g., do trait-level differences predict similar
expectation shifts across domains?) tractable without recollecting
data.

## The data

```{r data}
library(funfield)
data(overtimeESJT)
lapply(overtimeESJT, function(x) if (is.data.frame(x)) dim(x) else length(x))
```

## Variables and survey wording

### The scenario

Each participant saw exactly one scenario, randomly assembled from eight
situational factors. The Qualtrics template below shows the stem;
`${...}` placeholders are filled in per respondent from the eight
factors shown in the table that follows:

```{r scenario-template, echo = FALSE, results = "asis"}
cat("> ", overtimeESJT$codebook$scenario$template, "\n")
```

```{r scenario-factors}
knitr::kable(
  overtimeESJT$codebook$scenario$factors,
  col.names = c("Factor", "Levels (semicolon-separated)"),
  caption = "The eight factors that varied across respondents"
)
```

### Expected outcomes

After reading the scenario, each respondent rated each of the two
candidate actions on the nine expected-outcome items below
(0 = Very Low … 1 = Very High):

```{r wording-table}
knitr::kable(
  overtimeESJT$codebook$outcomes[, c("var", "label")],
  col.names = c("Variable", "Survey wording"),
  caption = "Outcome items, rated on [0, 1]"
)
```

The focal **predictor** is `Overtime` — the action coding
(0 = decline, 1 = agree to work overtime). The focal **outcome** is
`Likelihood`. The nine **mediators** are the expected-outcome ratings.

## Data preparation

Within-person deviate the Level-1 columns.

```{r prep}
L1cols <- c("Overtime","RelationshipQuality","EmployerValuation",
            "OutsideRelationships","Drained","CompanyHealth","Finances",
            "Appropriate","RaisePromotion","Punished","Likelihood")
mediators <- c("RelationshipQuality","EmployerValuation","OutsideRelationships",
               "Drained","CompanyHealth","Finances","Appropriate",
               "RaisePromotion","Punished")

dat <- overtimeESJT$PSI
for (v in L1cols) {
  dat[[v]] <- ave(dat[[v]], dat$p, FUN = function(x) x - mean(x))
}
```

## Normative model (no moderator)

```{r normative}
norm <- pathXMY(dat, X = "Overtime", Y = "Likelihood", M = mediators)
```

### Expectation paths (B1_MX): what does agreeing to overtime produce?

```{r bmx}
knitr::kable(
  subset(norm$tidy_loop, param == "B1_MX")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Expectation paths (Overtime -> M)"
)
```

Agreeing to work overtime is expected to substantially increase
`Drained` (+0.32), `RelationshipQuality` with the asker (+0.28),
`RaisePromotion` (+0.21), and `EmployerValuation` (+0.16). It is
expected to reduce `OutsideRelationships` (−0.16) and the
likelihood of being `Punished` by the company (−0.16). Smaller
positive expectations attach to `CompanyHealth`, `Finances`, and
`Appropriate`.

### Valuation paths (B1_YM): which expectations drive action likelihood?

```{r bym}
knitr::kable(
  subset(norm$tidy_loop, param == "B1_YM")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Valuation paths (M -> Likelihood)"
)
```

`Appropriate` (+0.90) and `Finances` (+0.89) are the strongest
positive drivers of taking the action. `RaisePromotion` (+0.65),
`EmployerValuation` (+0.60), and `RelationshipQuality` (+0.57) also
strongly increase likelihood. `Drained` (−0.33) and `Punished`
(−0.35) suppress likelihood. The valuation pattern is intuitive:
financial, evaluative, and relational gains pull respondents toward
the action; expected fatigue and punishment push them away.

### Indirect effects (B1_MX × B1_YM)

```{r bind}
knitr::kable(
  subset(norm$tidy_loop, param == "B1_MX * B1_YM")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Indirect effects through each mediator"
)
```

The pull toward agreeing comes mainly through expected gains in
`RelationshipQuality` (+0.16), `RaisePromotion` (+0.14), `Finances`
(+0.11), and `EmployerValuation` (+0.10). The push toward declining
comes through expected `Drained` (−0.10) and harm to
`OutsideRelationships` (−0.06).

## Situation-level moderation: is overtime paid?

`WO.cashReward` was randomly assigned at the person level: either the
company pays for the overtime work, or it doesn't. Code as a binary
`paid` indicator and merge:

```{r prep-paid}
cot <- overtimeESJT$cond[, c("p","cashReward")]
cot$paid <- as.numeric(cot$cashReward ==
  "the company will pay you overtime (above your normal pay rate) to work on the project")
dat <- merge(dat, cot[, c("p","paid")], by = "p")

mod_s <- pathXMY(dat, X = "Overtime", Y = "Likelihood",
                 M = mediators, Z = "paid", Z.within = FALSE)
```

### Does pay status moderate the expectation arm?

```{r bzmx-paid}
knitr::kable(
  subset(mod_s$tidy_loop, param == "BZ_MX")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Moderation of expectation paths by pay status (BZ_MX)"
)
```

Six of the nine expectation paths are significantly moderated by pay
status. When the company offers overtime pay, agreeing to overtime
is expected to produce substantially more `Finances` benefit
(+0.121) and to be substantially more `Appropriate` (+0.141), with
smaller positive shifts in `EmployerValuation`, `RaisePromotion`,
`RelationshipQuality`, and (less obviously) `OutsideRelationships`.
Pay status does not change expected `Drained`, `Punished`, or
`CompanyHealth` shifts — the cost side of the ledger is unchanged.

### Does that carry through to the indirect effect?

```{r bz-indirect-paid}
knitr::kable(
  subset(mod_s$tidy_loop, param == "BZ_MX * B1_YM")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Moderation of indirect effects by pay status (BZ_MX x B1_YM)"
)
```

When the action is paid, the indirect pull toward agreeing
strengthens through `Finances` and `Appropriate` (its two strongest
valuation paths) and through several smaller pathways. The increased
appeal of paid overtime comes from a sharper expected payoff on both
financial and appropriateness grounds, not from a shrinking of
expected costs.

### Decomposition and field view

```{r decompose-paid}
dec_paid <- pathXMY_decompose(dat, X = "Overtime", Y = "Likelihood",
                              M = mediators, Z = "paid", Z.within = FALSE)
knitr::kable(dec_paid$total, digits = 3,
             caption = "Total moderation BZ_YX[1] for pay status")
```

```{r plot-paid, fig.width = 10, fig.height = 7}
plotPathXMY(dec_paid,
            X_label = "Overtime", Y_label = "Likelihood",
            Z_label = "paid",
            title = "Overtime -> M -> Likelihood (Z = paid)")
```

The fan diagram's mediator arms are the per-mediator *loop* fits
(one X to M to Y model per mediator, overlaid). The straight
**X to Likelihood arrow** is the residual direct path from the
*joint* multi-mediator fit (added automatically by
`pathXMY(joint = TRUE)`, the default) — i.e. the
`Overtime -> Likelihood` effect after controlling for the *full*
set of nine mediators simultaneously. See the speeding vignette
for an extended discussion of why the loop coefficients (not the
per-mediator joint ones) are the right tool for inference about
a single mediator's role, and why `BZ_YX_joint` is best treated
as a system-level diagnostic (Wood, Adanu, & Harms, 2025).

```{r plot-paid-widget, eval = knitr::is_html_output()}
plotPathXMY_widget(dec_paid,
                   X_label = "Overtime", Y_label = "Likelihood",
                   Z_label = "paid",
                   Z_levels = c(-1, 1),
                   format   = "svg")
```

`paid` is binary (no / yes) and z-standardized inside `pathXMY()`,
so `Z = -1` and `Z = +1` map exactly to "no pay offered" and "pay
offered." The widget uses just those two frames. Toggling between
them: the upper arms (`Finances`, `Appropriate`,
`EmployerValuation`, `RaisePromotion`) all thicken when paid, while
the cost edges (`Drained`, `Punished`) stay fixed.

## Person-level moderation: vTime

`vTime` is the single-item self-report *"I value my time"* (SR_12 in
`$traits`, rescaled to [0, 1]). As a between-person trait, it uses
`Z.within = FALSE`.

```{r prep-vtime}
tr <- overtimeESJT$traits[, c("p","SR_12")]
names(tr)[2] <- "vTime"
tr <- tr[!is.na(tr$vTime), ]
dat2 <- merge(dat, tr, by = "p")

mod_p <- pathXMY(dat2, X = "Overtime", Y = "Likelihood",
                 M = mediators, Z = "vTime", Z.within = FALSE)
```

### Does vTime moderate the expectation arm?

```{r bzmx-vtime}
knitr::kable(
  subset(mod_p$tidy_loop, param == "BZ_MX")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Moderation of expectation paths by vTime (BZ_MX)"
)
```

Respondents who report valuing their time more expect agreeing to
overtime to produce *larger* gains in `EmployerValuation`,
`CompanyHealth`, `Finances`, and `RaisePromotion`. Each effect is
modest but the pattern is coherent: high-vTime respondents view
overtime as a more profitable investment of their time, both
financially and reputationally.

### Does that carry through to the indirect effect?

```{r bz-indirect-vtime}
knitr::kable(
  subset(mod_p$tidy_loop, param == "BZ_MX * B1_YM")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Moderation of indirect effects by vTime (BZ_MX x B1_YM)"
)
```

The corresponding indirect-path moderation is in the same direction
for the same mediators, though only the largest ones reach
significance after the multiplication by the valuation weights.

### Decomposition and field view

```{r decompose-vtime}
dec_vt <- pathXMY_decompose(dat2, X = "Overtime", Y = "Likelihood",
                            M = mediators, Z = "vTime", Z.within = FALSE)
knitr::kable(dec_vt$total, digits = 3,
             caption = "Total moderation BZ_YX[1] for vTime")
```

```{r plot-vtime, fig.width = 7, fig.height = 3.5}
plotPathXMY(dec_vt, mediator = "Finances",
            X_label = "Overtime", Y_label = "Likelihood",
            Z_label = "vTime",
            title = "Overtime -> Finances -> Likelihood (Z = vTime)")
```

```{r plot-vtime-widget, eval = knitr::is_html_output()}
plotPathXMY_widget(dec_vt, mediator = "Finances",
                   X_label = "Overtime", Y_label = "Likelihood",
                   Z_label = "vTime",
                   format  = "svg")
```

The vTime moderation is subtle but consistent: as time-valuing
rises, the `Overtime -> Finances` edge slightly thickens (people
who value time more expect a larger financial return from
agreeing). The other paths in the triangle are essentially fixed.

## Cross-scenario linkage

`overtimeESJT$traits` and `speedingESJT$traits` carry the same 30
self-report items for the same set of respondents. Cross-domain
analyses can be done by joining on `p`:

```{r cross-scenario, eval = FALSE}
data(speedingESJT)
both <- merge(
  overtimeESJT$traits[, c("p","SR_12")],
  speedingESJT$traits[, c("p","SR_12")],
  by = "p"
)  # identical SR_12 within p, by design — confirms the same-respondents claim
```

## Function reference

See `?pathXMY`, `?pathXMY_decompose`, `?plotPathXMY`,
`?plotPathXMY_widget`, `?esjtFromQualtrics`, and `?overtimeESJT`.
