---
title: "Moderated mediation with `pathXMY()`: the speeding EXSJT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Moderated mediation with pathXMY(): the speeding EXSJT}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png",
  dpi = 192,
  out.width = "70%"
)
options(width = 120)
```

This vignette demonstrates moderated mediation with `pathXMY()` using
the `speedingESJT` dataset — an **EXSJT** (Elaborated & eXperimental
Situational Judgment Test) in which a single driving scenario was
randomly varied across six factors, and each participant rated three
speed-choice actions on eight expected outcomes. Two moderators are
examined: **time pressure** (a situation-level factor randomly assigned
to each respondent) and **self-reported fast-driver tendency** (a
person-level trait).

This contrasts with the design of `coworkerESJT`, in which each
participant rated 12 distinct scenarios. There, situational factors vary
*within* person across scenarios. In the speeding EXSJT each participant
sees exactly one fully crossed scenario, so the six situational factors
vary *between* persons. All moderators in this vignette therefore use
`Z.within = FALSE`.

## The data

```{r data}
library(funfield)
data(speedingESJT)
lapply(speedingESJT, function(x) if (is.data.frame(x)) dim(x) else length(x))
```

The dataset is a named list with five components:

- **`$PSI`** — 999 rows: 333 persons × 3 actions, with 8
  expected-outcome ratings, the action's `Speed` coding, and the
  response `Likelihood`.
- **`$cond`** — the six randomly assigned situation factors per person,
  plus the assembled scenario text as displayed.
- **`$sit`** — situation perceptions (`Realistic`, `Relatable`) per
  person.
- **`$traits`** — 30 raw self-report items (`SR_01` … `SR_30`) plus
  demographics, per person.
- **`$codebook`** — survey wording, scale anchors, factor levels, and
  composite-scoring recipes.

All Likert ratings are stored on a [0, 1] proportion-of-maximum-possible
scale, so that within-person differences between two actions fall on
[-1, 1] — the natural metric for expectancy-value reasoning.

## Variables and survey wording

### The scenario

Each participant saw exactly one scenario, randomly assembled from six
situational factors. The Qualtrics scenario template lives at
`speedingESJT$codebook$scenario$template`; what that returns is shown
below in a readable format. The `${...}` placeholders are filled in per
respondent from the six factors that follow:

```{r scenario-template, echo = FALSE, results = "asis"}
cat("> ", speedingESJT$codebook$scenario$template, "\n")
```

```{r scenario-factors}
knitr::kable(
  speedingESJT$codebook$scenario$factors,
  col.names = c("Factor", "Levels (semicolon-separated)"),
  caption = "The six factors that varied across respondents"
)
```

Each respondent's fully assembled scenario is in
`speedingESJT$cond$scenario`. Below are the first three (i.e., what
`speedingESJT$cond$scenario[1:3]` returns, again formatted for
readability):

```{r scenario-assembled, echo = FALSE, results = "asis"}
ss <- speedingESJT$cond$scenario[1:3]
for (i in seq_along(ss)) {
  cat("> **Respondent ", i, ".** ", ss[i], "\n\n", sep = "")
}
```

Notice how the same template produces materially different scenarios
through random factor assignment. The factor levels that were combined
to produce each of the three scenarios above:

```{r scenario-cond}
knitr::kable(
  speedingESJT$cond[1:3, c("vehicle","weather","traffic",
                           "timePressure","context","policeFreq")],
  caption = "Factor levels for respondents 1-3"
)
```

### Expected outcomes

After reading the scenario, each respondent rated each of the three
candidate actions on the eight expected-outcome items below (0 = Very
Low … 1 = Very High):

```{r wording-table}
knitr::kable(
  speedingESJT$codebook$outcomes[, c("var", "label")],
  col.names = c("Variable", "Survey wording"),
  caption = "Outcome items (PSI), rated on [0, 1]"
)
```

The focal **predictor** is `Speed` — the action's encoded speed level (0
= at the limit, 0.5 = +5 mph, 1 = +10+ mph). The focal **outcome** is
`Likelihood` — how likely the respondent says they would take the
action. The eight **mediators** are the expected-outcome ratings.

## Data preparation

Within-person deviate all Level-1 columns. This is the only required
preprocessing — situation factors (in `$cond`) and traits (in `$traits`)
are already at the appropriate level and need no deviation.

```{r prep}
L1cols <- c("Speed","MoneyCost","FunDrive","IntQuality","Appropriate",
            "OnTime","Crash","Injured","Ticket","Likelihood")
mediators <- c("MoneyCost","FunDrive","IntQuality","Appropriate",
               "OnTime","Crash","Injured","Ticket")

dat <- speedingESJT$PSI
for (v in L1cols) {
  dat[[v]] <- ave(dat[[v]], dat$p, FUN = function(x) x - mean(x))
}
```

## Normative model (no moderator)

How does each expected outcome mediate the link between an action's
speed and how likely the respondent is to take it, averaging across
people and conditions?

```{r normative}
norm <- pathXMY(dat, X = "Speed", Y = "Likelihood", M = mediators)

## Loop pass: one X -> M -> Y regression PER MEDIATOR, fit independently.
## A B1_YM row here is the slope of Likelihood on that mediator
## controlling only for Speed (no other mediators in the equation).
norm$tidy_loop

## Joint pass: a SINGLE simultaneous regression with all eight mediators
## together in the Y equation. B1_YM_joint slopes are partial slopes net
## of the other seven mediators -- not comparable to the loop B1_YM above.
norm$tidy_joint
```

### Expectation and valuation paths, side by side

Each mediator sits on two paths. `B1_MX` is the **expectation** path —
how much an action's speed shifts that expected outcome. `B1_YM` is the
**valuation** path — how much that expected outcome, in turn, drives
action likelihood. `pathXMY_pairtable()` reshapes the long `tidy_loop`
table so the two paths sit next to each other, one row per mediator, and
the `mediator` column is not repeated. `group_kable()` then prints that
table with each parameter block set under its own spanning header.

```{r b1-pair}
group_kable(
  pathXMY_pairtable(norm, c("B1_MX", "B1_YM")),
  groups = c(" " = 1,
             "Expectation paths (B1_MX: Speed &rarr; M)"      = 4,
             "Valuation paths (B1_YM: M &rarr; Likelihood)"   = 4),
  col_labels = c("Mediator", "est", "se", "z", "p",
                             "est", "se", "z", "p")
)
```

*Expectation paths (`B1_MX`).* Faster speeds are expected to
substantially increase `Ticket`, `Crash`, `Injured`, `OnTime`, and (more
modestly) `MoneyCost`, and to substantially reduce perceived
appropriateness (`Appropriate`). `FunDrive` and `IntQuality` are
essentially uncorrelated with `Speed` at the within-person level —
perceived enjoyment of the drive and quality of the next meeting do not
move much with the action's speed.

*Valuation paths (`B1_YM`).* `FunDrive` and `Appropriate` are the
strongest positive drivers of whether the respondent says they'd take
the action, followed by `IntQuality` and `OnTime`. `Crash`, `Injured`,
`Ticket`, and `MoneyCost` all suppress likelihood. The signs are
intuitive throughout.

### Indirect effects (B1_MX × B1_YM)

```{r bind}
knitr::kable(
  subset(norm$tidy_loop, param == "B1_MX * B1_YM")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Indirect effects through each mediator"
)
```

The pathways through `Ticket`, `Appropriate`, `Crash`, and `Injured` are
all large and negative: faster speeds are expected to produce tickets
and accidents (each of which strongly suppresses action likelihood) and
to be less appropriate (which also strongly suppresses likelihood). The
`OnTime` pathway is the lone strong positive: faster speeds are expected
to get the respondent to the meeting on time, which increases
likelihood. `FunDrive` and `IntQuality` contribute essentially nothing
here — because faster speeds don't shift those expectations, the
pathways are near zero despite their strong valuation paths.

## Total moderation and its decomposition {#total-moderation-and-its-decomposition}

### (here: What personal and situational factors affect the likelihood of speeding, and why?)

The `pathXMY()` function is particularly useful for understanding the
personal and situational factors that may affect an `X → Y` pathway, and
why they may do so from within an expectancy×value lens. The full identity is

$$\beta^*_{Z_{YX}} \approx\; \beta_{Z_{MX}} \cdot \beta_{1_{YM}}
\;+\; \beta_{1_{MX}} \cdot \beta_{Z_{YM}}
\;+\; \beta_{Z_{YX}}$$

When `X` is an action, `M` is an outcome, and `Y` is the rated action likelihood, the model can account for behavior in the manner below:
- **Term 0** (`BZ_YX`): This value, on the left-hand-side of the equation, indicates the total degree to which the variable `X` moderates the likelihood of doing the action (here: speeding).
- **Term 1** (`BZ_MX × B1_YM`): `Z` shifts *the degree to which `X` is expected to affect `M`* — the **expectation** route, which affects the likelihood of doing the action if `M` tends to be valued.
- **Term 2** (`B1_MX × BZ_YM`): `Z` shifts *how much outcome `M` affects `Y`* — the **valuation** route, which effects the action likelihood if `X` tends to affect `M`.
- **Term 3** (`BZ_YX`): the residual moderation of the `X → Y` pathway by variable `Z`, after accounting for this mediator. (Here: concerns extent to which the person/situation factor's affect on speeding can't be accounted for by this mediator.)

Note also that the decomposition will result in an *approximation* of the initial `BZ_YX` term prior to considering the mediator.  It will not be exactly equal for reasons detailed by Muller, Judd, & Yzerbyt, (2005) - although the decomposition will tend to be very close.

When a moderator produces a large (`BZ_MX × B1_YM`) or (`B1_MX × BZ_YM`) effect, this can be described as formally indicating a *reason* for the behavior. 

## Example of Person-level moderation: self-reported fast-driver tendency

`SRFastDriver` is the [0, 1] rescaled response to *"I regularly drive
far over the speed limit when driving"* — item `SR_30` in `$traits`. As
a between-person trait, it uses `Z.within = FALSE`. Some respondents are
missing this item:

```{r prep-srfast}
sr <- speedingESJT$traits[, c("p","SR_30")]
sr <- sr[!is.na(sr$SR_30), ]
names(sr)[2] <- "SRFastDriver"
dat2 <- merge(dat, sr, by = "p")
cat("Persons with SRFastDriver:", length(unique(dat2$p)), "\n")

mod_p <- pathXMY(dat2, X = "Speed", Y = "Likelihood",
                 M = mediators, Z = "SRFastDriver", Z.within = FALSE)
```

### The total moderation:

`pathXMY_decompose()` fits both the no-mediator model (giving just
`BZ_YX`) and the full mediated model (giving the three components) and tabulates them per mediator.

```{r decompose-srfast}
dec <- pathXMY_decompose(dat2, X = "Speed", Y = "Likelihood",
                         M = mediators, Z = "SRFastDriver",
                         Z.within = FALSE)
```



```{r decompose-total}
knitr::kable(dec$total, digits = 3,
             caption = "Total moderation BZ_YX[1] (no mediator)")
```

Self-reported fast-driver tendency moderates the `Speed → Likelihood`
link by **+0.25** — a substantial effect: respondents one SD higher on
`SRFastDriver` reported increasing their likelihood of speeding by 25% (i.e., an extra 0.25 units
on the [0, 1] outcome) when shifting from at-limit to +10+ mph speeds.

### Decomposition per mediator

```{r decompose-table}
knitr::kable(dec$components, digits = 3,
             caption = "Decomposition of BZ_YX[1] through each single mediator")
```

Each mediator's three components plus residual sum to \~0.25 — that's
the algebraic identity at work. The `sum (1+2+3)` row should equal
`BZ_YX[1]` exactly in the linear case (small deviations come from
numerical precision).

**Where does the moderation route?** The largest single-mediator
expectation-route is through `Appropriate`: `BZ_MX × B1_YM = 0.073`.
Term 2 (valuation route) is negligible for every mediator here — fast
drivers don't differentially *weight* outcomes, they have different
*expectations*. And **most of `BZ_YX[1]` sits in the residual `BZ_YX[2]`
column** (≈ 0.18–0.24 for every mediator). That isn't a failure of the
decomposition — it's the signature of moderation routed through
*multiple mediators jointly*. With a single mediator in the model, every
other mediator's contribution gets absorbed into the residual.

### Can we identify likely reasons that regular speeders tend to speed?

```{r bz-pair-srfast}
group_kable(
  pathXMY_pairtable(mod_p, c("BZ_MX", "BZ_YM")),
  groups = c(" " = 1,
             "Fast-driver moderation: expectation arm (BZ_MX)" = 4,
             "Fast-driver moderation: valuation arm (BZ_YM)"   = 4),
  col_labels = c("Mediator", "est", "se", "z", "p",
                             "est", "se", "z", "p")
)
```

On the **expectation** side (`BZ_MX`), self-described fast drivers expect
speeding to be less catastrophic — smaller `Crash`, `Injured`, `Ticket`,
and `MoneyCost` shifts (all *p* \< .01) — more enjoyable (`FunDrive`,
*p* \< .001), and substantially more role-appropriate (`Appropriate`,
*p* \< .001, by far the largest effect). Six of the eight expectation
paths are significantly moderated.

On the **valuation** side (`BZ_YM`), *not one* of the eight paths
reaches *p* \< .05; the nearest is `OnTime` (*b* = +0.12, *p* = .11).
Fast drivers do not differentially *weight* the expected outcomes — they
hold different *expectations* about what speeding produces. Taken with
the time-pressure result, the pattern is consistent: across both a
situational pressure and a person-level trait, every significant
moderation effect lives in the expectation arm. Both kinds of moderator
seem to operate by changing what people expect of speeding, not by
changing how they value those expectations — at least at the resolution
these data afford. The one place a genuine valuation-arm effect *does*
surface — `DrivingSkill` moderating the weight placed on `Crash` — is
taken up below under [When does Term 2 carry the
moderation?](#when-does-term-2-carry-the-moderation).

### Does these differing expectancies & values create reasons for speeding?

As before, the two columns are the model's two **routes** by which the
moderator reshapes the overall `Speed → Likelihood` link — the
**expectation route** (`BZ_MX × B1_YM`) and the **valuation route**
(`B1_MX × BZ_YM`). Significant entries are this model's formal account
of *why* self-described fast drivers are more likely to speed: the
[decomposition section](#total-moderation-and-its-decomposition) below
shows that fast-driver tendency moderates the direct
`Speed → Likelihood` path by `BZ_YX[1]` **= +0.25**, and these
per-mediator route effects are the pieces that sum toward it.

```{r bz-indirect-srfast}
group_kable(
  pathXMY_pairtable(mod_p, c("BZ_MX * B1_YM", "B1_MX * BZ_YM")),
  groups = c(" " = 1,
             "Expectation route (BZ_MX &times; B1_YM)" = 4,
             "Valuation route (B1_MX &times; BZ_YM)"   = 4),
  col_labels = c("Mediator", "est", "se", "z", "p",
                             "est", "se", "z", "p")
)
```

The expectation route does essentially all the work. Fast-driver
tendency strengthens the indirect pull toward the faster action through
`Appropriate` (*p* \< .001, the largest single route), `FunDrive`,
`Crash`, `Injured`, and `Ticket` (all *p* \< .005) and `MoneyCost` (*p*
= .03) — six significant expectation-route reasons. The valuation route
contributes nothing significant (nearest: `OnTime`, *p* = .11). So the
*reason* fast drivers speed more is, item by item, that they expect
speeding to be more appropriate and less costly — not that they weight
those outcomes differently. The contrast with the time-pressure result
is informative: time pressure shifts the *upside* (OnTime, IntQuality);
fast-driver tendency shifts the *downside* (the expected costs of
speeding feel smaller, and the action feels more appropriate). In both
cases the moderation routes almost entirely through shifted
expectations.

### Field-style diagrams

`plotPathXMY()` renders the X-M-Y structure in the funfield
visualization style:

- Nodes are colored by their **expected score when X is held at 1** —
  positive in blue, negative in red, gradient by magnitude. X is by
  definition full blue (score = 1); each mediator is colored by its
  `B1_MX` coefficient; Y is colored by the implied total effect at X=1.
- Y (Likelihood) is drawn as a left-facing triangle to signal that it's
  the focal action node.
- Edges are **dashed for negative**, **solid for positive**, with
  thickness scaled by path magnitude (`negDashed = TRUE`,
  `maximum = 0.8`). Edge labels show the full `b1 + bZ(Z)` form.

**Single-mediator triangle** for the dominant route through
`Appropriate`:

```{r plot-appropriate, fig.width = 7, fig.height = 3.5}
plotPathXMY(dec, mediator = "Appropriate",
            X_label = "Speed", Y_label = "Likelihood",
            Z_label = "SRFast",
            title = "Speed -> Appropriate -> Likelihood (Z = SRFastDriver)")
```

Reading the diagram: a one-unit increase in `Speed` lowers `Appropriate`
by 0.45 *on average* (dashed edge, negative), but among fast drivers
(high `SRFast`) that drop is dampened by +0.14 — they perceive faster
driving as less of an appropriateness violation. `Appropriate` strongly
drives `Likelihood` (+0.53, solid) and that valuation is *not* moderated
by `SRFast` (≈ +0.00). The residual `BZ_YX[2]` direct path (+0.18)
reflects the joint contribution of the seven other mediators not shown
here. The Y node renders red because the total predicted Y at
`Speed = 1` is negative — speeding produces a net-negative expected
likelihood when only `Appropriate` mediates.

**Full multi-mediator fan view** — pass no `mediator` argument and all
available mediators are arranged in a fan between X and Y:

```{r plot-fan, fig.width = 10, fig.height = 7}
plotPathXMY(dec,
            X_label = "Speed", Y_label = "Likelihood",
            Z_label = "SRFast",
            title = "Full X-M-Y field for speeding decisions")
```

The fan diagram is a **summary overlay**, not a single fit. The mediator
arms are read off the per-mediator *loop* fits — one X to M to Y model
per mediator — so each `Speed -> M_k` and `M_k -> Likelihood` arm is
interpretable on its own without worrying about how the other mediators
are competing for variance. The straight **X to Likelihood arrow** is
different: it is drawn from the *joint* multi-mediator fit (added
automatically by `pathXMY(joint = TRUE)`, the default), and shows the
residual direct effect of `Speed` after controlling for the *full*
mediator set.

The mediator column reveals which expectations move with Speed: `Ticket`
(+0.55) and `Appropriate` (-0.45) are the strongest in opposite
directions; `FunDrive` and `IntQuality` are essentially unaffected by
Speed (near-white). The right column of edges shows the valuation
weights (`B1_YM`). Edge thickness in the fan diagnoses where the
indirect-effect mass concentrates.

### Joint multi-mediator fit: residual direct path and stability

The same `pathXMY()` call that produced the loop coefficients also
returns simultaneous-fit parameters under the `_joint` suffix. The most
useful of these are the residual direct path and its Z-moderation:

```{r joint-direct}
joint_yx <- subset(mod_p$tidy_joint, param %in% c("B1_YX_joint","BZ_YX_joint"))
joint_yx$mediator <- NULL
knitr::kable(joint_yx, digits = 3,
  caption = "Residual direct path X -> Y from the joint fit of all 8 mediators (Z = SRFastDriver)")
```

`B1_YX_joint` is the residual `Speed -> Likelihood` path controlling for
all eight expected outcomes simultaneously. If the measured mediator set
has captured the within-person mechanisms, it should be near zero.
`BZ_YX_joint` is the analogous diagnostic for moderation: an
`SRFastDriver` effect on `Speed -> Likelihood` that does *not* run
through any of the eight outcomes shows up here. The decomposition table
earlier in this section flagged `BZ_YX[2]` as substantial (≈ 0.18-0.24)
for every single-mediator fit; the joint result tells us whether the
mediator set collectively absorbs that moderation.

**Why prefer loop coefficients for per-mediator inference.** The joint
fit also returns `B1_YM_joint` and `BZ_YM_joint` rows. When mediators
are numerous and correlated — as the eight expected outcomes here
certainly are — the M-by-Z product terms used to estimate `BZ_YM_joint`
are highly collinear with each other, which inflates the standard errors
substantially compared to the single-mediator loop fit (Wood, Adanu, &
Harms, 2025). The loop estimates are not contaminated by this
multicollinearity because each loop fit estimates only one mediator's
`BZ_YM` at a time. The asymmetry can be made explicit:

```{r joint-stability}
loop <- subset(mod_p$tidy_loop,
               param == "BZ_YM" & !is.na(mediator))[, c("mediator","est","se")]
join <- subset(mod_p$tidy_joint,
               param == "BZ_YM_joint")[, c("mediator","est","se")]
names(loop)[2:3] <- c("loop_est","loop_se")
names(join)[2:3] <- c("joint_est","joint_se")
cmp <- merge(loop, join, by = "mediator")
cmp$se_inflation <- cmp$joint_se / cmp$loop_se
knitr::kable(cmp, digits = 3,
  caption = "BZ_YM (valuation moderation) per mediator: loop vs joint, with SE inflation factor")
```

Joint SEs are inflated compared to the loop SEs (see the `se_inflation`
column), and the joint point estimates jitter substantially across
mediators when the mediator set is large and correlated. For inference
about *which* mediators carry the valuation-route moderation, the loop
coefficient is the right tool. The joint fit is best treated as a
system-level diagnostic — most useful for `B1_YX_joint` / `BZ_YX_joint`
(which are stable, because they estimate a single residual path), less
useful for the per-mediator joint coefficients.

> Wood, D., Adanu, E. K., & Harms, P. D. (2025). Expectancy x value
> models of the relations between demographic, psychological, and
> situational factors and speeding behavior. *Journal of Safety
> Research*, **93**, 135-147. <doi:10.1016/j.jsr.2025.02.012>

### Two-panel low-Z vs high-Z view

`plotPathXMY_ZLH()` draws the field at a low and high value of `Z`
side-by-side, with coefficients collapsed to their effective value at
that `Z`. This makes the moderation visually concrete: edge widths thin
out, node colors fade, and the implied total effect on `Y` shifts
between panels.

```{r plot-zlh-fan, fig.width = 12, fig.height = 6}
plotPathXMY_ZLH(dec,
                X_label = "Speed", Y_label = "Likelihood",
                Z_label = "SRFast",
                Z_levels = c(-1, 1))
```

At `SRFast = -1` (low fast-driver tendency) the expected costs of
speeding are maximal: large red `Appropriate` node, thick negative edges
from `Crash`/`Injured`/`Ticket`, and the `Likelihood` triangle glows
deep red — the field predicts a substantial *decrease* in action
likelihood as `Speed` rises.

At `SRFast = +1` the same field has visibly attenuated: every cost
mediator's Speed-loading shrinks (smaller numbers, thinner edges), the
`Appropriate` node lightens, and the `Likelihood` triangle fades — the
predicted decrease is much smaller. Self-reported fast drivers expect
speeding to be less costly across the board, and the field-level summary
captures all of those shifts at once.

### Toggleable Z-scrubbing widget

The side-by-side panels make the contrast easy to compare, but
overlaying them in place and toggling makes the moderation feel even
sharper — you can watch the edge thicknesses and node colors swap as you
click forward / back. `plotPathXMY_widget()` renders one frame per Z
value and embeds them in a self-contained HTML widget with Back /
Forward buttons.

```{r plot-widget, eval = knitr::is_html_output()}
plotPathXMY_widget(dec,
                   X_label = "Speed", Y_label = "Likelihood",
                   Z_label = "SRFast",
                   format  = "svg")
```

The default `Z_levels = c(-1, 0, 1)` toggles between one SD below the
mean, the mean, and one SD above. The widget renders in HTML output
only; in PDF the chunk is skipped. Path coefficients are linear in `Z`
so frames between the endpoints are trivially interpolatable; values at
`|Z| > 1` are often outside the well-supported range of the data, so are
not in the default — pass an explicit `Z_levels` vector if you want a
wider range.

### When does Term 2 carry the moderation? {#when-does-term-2-carry-the-moderation}

A counter-example shows valuation routing in isolation. Consider
`DrivingSkill` (SR_07: *"I would say that I am a skilled driver"*) as
the moderator, with `Crash` as the focal mediator:

```{r decompose-skill}
tr <- speedingESJT$traits[, c("p","SR_07")]
names(tr)[2] <- "DrivingSkill"
tr <- tr[!is.na(tr$DrivingSkill), ]
dat_sk <- merge(dat, tr, by = "p")

dec_sk <- pathXMY_decompose(dat_sk, X = "Speed", Y = "Likelihood",
                            M = mediators, Z = "DrivingSkill",
                            Z.within = FALSE)
knitr::kable(
  subset(dec_sk$components, mediator == "Crash"),
  digits = 3,
  caption = "Crash decomposition under Z = DrivingSkill"
)
```

Here Term 1 (`BZ_MX × B1_YM = -0.002`) is essentially zero — self-rated
skilled drivers don't expect a *different* crash likelihood from
speeding. But Term 2 (`B1_MX × BZ_YM = -0.061`) is significantly
non-zero: skilled-driver respondents weight `Crash` *more* heavily as a
deterrent than unskilled-driver respondents do. In the decomposition's
vocabulary this is a *reason* located in the **valuation route**: to
whatever degree skilled drivers hold back from speeding, part of the
account is that a potential crash simply weighs more heavily for them —
not that they expect crashes any differently. Without
`pathXMY_decompose()`, this valuation-route reason would be invisible in
the standard `BZ_MX` table.

```{r plot-skill, fig.width = 7, fig.height = 3.5}
plotPathXMY(dec_sk, mediator = "Crash",
            X_label = "Speed", Y_label = "Likelihood",
            Z_label = "Skill",
            title = "Speed -> Crash -> Likelihood (Z = DrivingSkill)")
```

Toggling through `Skill = -1, 0, +1` makes the valuation-only moderation
pattern especially clear: the `Speed -> Crash` edge barely budges (BZ_MX
is null), while the `Crash -> Likelihood` edge steepens as `Skill` rises
— and the `Likelihood` node correspondingly deepens in red. The whole
change rides on a single arm.

```{r plot-skill-widget, eval = knitr::is_html_output()}
plotPathXMY_widget(dec_sk, mediator = "Crash",
                   X_label = "Speed", Y_label = "Likelihood",
                   Z_label = "Skill",
                   format  = "svg")
```


## Situation-level moderation: time pressure

The decomposition framework applies unchanged to *situational*
moderators — they are handled statistically exactly as the personal
moderator above; only the substantive reading differs.

`timePressure` was randomly assigned at the person level — each
respondent saw exactly one of three levels
(`comfortably ahead of schedule`, `a bit late`, `quite late`). Code it
as a continuous "lateness" variable (0/0.5/1), merge it, and fit both
the moderated model and its decomposition:

```{r prep-timepressure}
late_code <- c(`comfortably ahead of schedule` = 0,
               `a bit late`                    = 0.5,
               `quite late`                    = 1)
dat$late <- late_code[speedingESJT$cond$timePressure[match(dat$p,
                                                           speedingESJT$cond$p)]]

mod_s <- pathXMY(dat, X = "Speed", Y = "Likelihood",
                 M = mediators, Z = "late", Z.within = FALSE)
dec_s <- pathXMY_decompose(dat, X = "Speed", Y = "Likelihood",
                           M = mediators, Z = "late", Z.within = FALSE)
```

### The total moderation

```{r decompose-total-late}
knitr::kable(dec_s$total, digits = 3,
             caption = "Total moderation BZ_YX[1] by time pressure (no mediator)")
```

Where the fast-driver trait moved the `Speed → Likelihood` link by a
substantial, significant **+0.25**, time pressure's total moderation is
only **+0.04**, and is *not* statistically significant (*p* = .18).
Read on its own, that says time pressure does not reliably change how
strongly faster speeds pull respondents toward speeding.

A non-significant total is exactly where the decomposition earns its
keep. It can mean the moderator is genuinely irrelevant — or that real
route-specific shifts are present but too small, or too offsetting, to
surface in the aggregate. Only the per-mediator breakdown tells the two
apart.

### Decomposition per mediator

```{r decompose-table-late}
knitr::kable(dec_s$components, digits = 3,
             caption = "Decomposition of the time-pressure moderation through each single mediator")
```

The decomposition localizes the modest signal. Two **expectation-route**
entries (Term 1, `BZ_MX × B1_YM`) stand out: time pressure strengthens
the pull toward speeding through the expected `OnTime` gain (+0.031,
*p* \< .001) and the expected `IntQuality` gain (+0.026, *p* = .002),
with `Appropriate` a borderline third. The **valuation route** (Term 2,
`B1_MX × BZ_YM`) is null for every mediator. The residual
`BZ_YX (direct)` column — the total minus that mediator's two routes —
adds nothing the total did not already say: with a near-null total
there is little for it to carry. So time pressure does produce a couple
of genuine, significant expectation-route *reasons* to speed — chiefly a
larger expected on-time payoff — but they are small and never
accumulate into a reliable aggregate effect.

### Which expectancies and values does time pressure shift?

```{r bz-pair-late}
group_kable(
  pathXMY_pairtable(mod_s, c("BZ_MX", "BZ_YM")),
  groups = c(" " = 1,
             "Time-pressure moderation: expectation arm (BZ_MX)" = 4,
             "Time-pressure moderation: valuation arm (BZ_YM)"   = 4),
  col_labels = c("Mediator", "est", "se", "z", "p",
                             "est", "se", "z", "p")
)
```

The raw arms tell the same story as the decomposition, one step
upstream. On the **expectation** side (`BZ_MX`), time pressure
significantly amplifies the expected on-time gain from speeding
(`OnTime`, *p* \< .001 — the largest effect) and the expected
meeting-quality gain (`IntQuality`, *p* \< .001), with appropriateness a
borderline third (`Appropriate`, *p* ≈ .05). Other expectations —
including crash and ticket expectations — are not moved. People under
time pressure don't expect *different consequences* from speeding; they
expect a *larger payoff* from getting there on time.

On the **valuation** side (`BZ_YM`), nothing reaches *p* \< .05: time
pressure does not detectably change how heavily any expected outcome is
weighted. The two nearest misses are both cost outcomes — `Ticket` (*b*
= -0.14, *p* = .06) and `MoneyCost` (*b* = -0.13, *p* = .10) — hinting
that respondents may weight the financial costs of speeding slightly
more heavily when rushed, but neither effect is significant. This echoes
the fast-driver case: **expectation-arm moderation is the common case in
these data; valuation-arm moderation is comparatively rare.**

### Do those shifts become reasons to speed?

The expectation-route (`BZ_MX × B1_YM`) and valuation-route
(`B1_MX × BZ_YM`) products turn the raw arm shifts above into the
per-mediator *reasons* of the decomposition — Term 1 and Term 2, with
their significance tests laid side by side:

```{r bz-indirect-late}
group_kable(
  pathXMY_pairtable(mod_s, c("BZ_MX * B1_YM", "B1_MX * BZ_YM")),
  groups = c(" " = 1,
             "Expectation route (BZ_MX &times; B1_YM)" = 4,
             "Valuation route (B1_MX &times; BZ_YM)"   = 4),
  col_labels = c("Mediator", "est", "se", "z", "p",
                             "est", "se", "z", "p")
)
```

The only reliable *reasons* are expectation-route ones: time pressure
strengthens the pull toward speeding through the expected `OnTime`
(*p* \< .001) and `IntQuality` (*p* = .002) gains, with `Appropriate`
borderline (*p* = .06). The valuation route yields nothing significant
(nearest: `Ticket`, *p* = .06). But unlike the fast-driver case — where
six expectation-route reasons summed to a large, significant **+0.25**
total — here the two genuine reasons are modest and, as the total
moderation already showed, do not accumulate into a statistically
reliable change in overall speeding propensity. The decomposition's
value in this example is exactly that: it pinpoints *which* expectations
time pressure shifts, and confirms those shifts are real, even though
the aggregate moderation is not itself significant.

## Scoring named trait composites

If you want the Big Five and other named scales from Wood, Harms, & Cho
(2023), use the recipes in `speedingESJT$codebook$trait_scales`:

```{r scoring, eval = FALSE}
tr <- speedingESJT$traits
tr$Extra <- rowMeans(cbind(1 - tr$SR_15, tr$SR_16, tr$SR_17), na.rm = TRUE)
tr$Agree <- rowMeans(cbind(tr$SR_18, 1 - tr$SR_19, tr$SR_20), na.rm = TRUE)
tr$Consc <- rowMeans(cbind(1 - tr$SR_21, 1 - tr$SR_22, tr$SR_23), na.rm = TRUE)
tr$Neuro <- rowMeans(cbind(tr$SR_24, tr$SR_25, 1 - tr$SR_26), na.rm = TRUE)
tr$Open  <- rowMeans(cbind(tr$SR_27, 1 - tr$SR_28, tr$SR_29), na.rm = TRUE)
```

## Function reference

See `?pathXMY`, `?pathXMY_pairtable`, `?group_kable`,
`?pathXMY_decompose`, `?plotPathXMY`, `?plotPathXMY_ZLH`,
`?plotPathXMY_widget`, and `?speedingESJT`.
