---
title: "Moderated mediation with `pathXMY()`: the speeding EXSJT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Moderated mediation with pathXMY(): the speeding EXSJT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette demonstrates moderated mediation with `pathXMY()` using the
`speedingESJT` dataset — an **EXSJT** (Elaborated & eXperimental Situational
Judgment Test) in which a single driving scenario was randomly varied across
six factors, and each participant rated three speed-choice actions on eight
expected outcomes. Two moderators are examined: **time pressure** (a
situation-level factor randomly assigned to each respondent) and
**self-reported fast-driver tendency** (a person-level trait).

This contrasts with the design of `coworkerESJT`, in which each participant
rated 12 distinct scenarios. There, situational factors vary *within* person
across scenarios. In the speeding EXSJT each participant sees exactly one
fully crossed scenario, so the six situational factors vary *between*
persons. All moderators in this vignette therefore use `Z.within = FALSE`.

## The data

```{r data}
library(funfield)
data(speedingESJT)
lapply(speedingESJT, function(x) if (is.data.frame(x)) dim(x) else length(x))
```

The dataset is a named list with five components:

- **`$PSI`** — 999 rows: 333 persons × 3 actions, with 8 expected-outcome
  ratings, the action's `Speed` coding, and the response `Likelihood`.
- **`$cond`** — the six randomly assigned situation factors per person,
  plus the assembled scenario text as displayed.
- **`$sit`** — situation perceptions (`Realistic`, `Relatable`) per person.
- **`$traits`** — 30 raw self-report items (`SR_01` … `SR_30`) plus
  demographics, per person.
- **`$codebook`** — survey wording, scale anchors, factor levels, and
  composite-scoring recipes.

All Likert ratings are stored on a [0, 1] proportion-of-maximum-possible
scale, so that within-person differences between two actions fall on
[-1, 1] — the natural metric for expectancy-value reasoning.

## Variables and survey wording

```{r wording-table}
knitr::kable(
  speedingESJT$codebook$outcomes[, c("var", "label")],
  col.names = c("Variable", "Survey wording"),
  caption = "Outcome items (PSI), rated on [0, 1]"
)
```

The focal **predictor** is `Speed` — the action's encoded speed level
(0 = at the limit, 0.5 = +5 mph, 1 = +10+ mph). The focal **outcome** is
`Likelihood` — how likely the respondent says they would take the action.
The eight **mediators** are the expected-outcome ratings.

## Data preparation

Within-person deviate all Level-1 columns. This is the only required
preprocessing — situation factors (in `$cond`) and traits (in `$traits`) are
already at the appropriate level and need no deviation.

```{r prep}
L1cols <- c("Speed","MoneyCost","FunDrive","IntQuality","Appropriate",
            "OnTime","Crash","Injured","Ticket","Likelihood")
mediators <- c("MoneyCost","FunDrive","IntQuality","Appropriate",
               "OnTime","Crash","Injured","Ticket")

dat <- speedingESJT$PSI
for (v in L1cols) {
  dat[[v]] <- ave(dat[[v]], dat$p, FUN = function(x) x - mean(x))
}
```

## Normative model (no moderator)

How does each expected outcome mediate the link between an action's speed
and how likely the respondent is to take it, averaging across people and
conditions?

```{r normative}
norm <- pathXMY(dat, X = "Speed", Y = "Likelihood", M = mediators)
norm$tidy
```

### Expectation paths (B1_MX): what is speeding expected to produce?

```{r bmx}
knitr::kable(
  subset(norm$tidy, param == "B1_MX")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Expectation paths (Speed -> M)"
)
```

Faster speeds are expected to substantially increase `Ticket`, `Crash`,
`Injured`, `OnTime`, and (more modestly) `MoneyCost`, and to substantially
reduce perceived appropriateness (`Appropriate`). `FunDrive` and
`IntQuality` are essentially uncorrelated with `Speed` at the within-person
level — perceived enjoyment of the drive and quality of the next meeting do
not move much with the action's speed.

### Valuation paths (B1_YM): which expected outcomes drive action likelihood?

```{r bym}
knitr::kable(
  subset(norm$tidy, param == "B1_YM")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Valuation paths (M -> Likelihood)"
)
```

`FunDrive` and `Appropriate` are the strongest positive drivers of whether
the respondent says they'd take the action, followed by `IntQuality` and
`OnTime`. `Crash`, `Injured`, `Ticket`, and `MoneyCost` all suppress
likelihood. The signs are intuitive throughout.

### Indirect effects (B1_MX × B1_YM)

```{r bind}
knitr::kable(
  subset(norm$tidy, param == "B1_MX * B1_YM")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Indirect effects through each mediator"
)
```

The pathways through `Ticket`, `Appropriate`, `Crash`, and `Injured` are
all large and negative: faster speeds are expected to produce tickets and
accidents (each of which strongly suppresses action likelihood) and to be
less appropriate (which also strongly suppresses likelihood). The `OnTime`
pathway is the lone strong positive: faster speeds are expected to get the
respondent to the meeting on time, which increases likelihood. `FunDrive`
and `IntQuality` contribute essentially nothing here — because faster
speeds don't shift those expectations, the pathways are near zero despite
their strong valuation paths.

## Situation-level moderation: time pressure

`timePressure` was randomly assigned at the person level — each respondent
saw exactly one of three levels (`comfortably ahead of schedule`,
`a bit late`, `quite late`). Code it as a continuous "lateness" variable
(0/0.5/1) and merge:

```{r prep-timepressure}
late_code <- c(`comfortably ahead of schedule` = 0,
               `a bit late`                    = 0.5,
               `quite late`                    = 1)
dat$late <- late_code[speedingESJT$cond$timePressure[match(dat$p,
                                                           speedingESJT$cond$p)]]

mod_s <- pathXMY(dat, X = "Speed", Y = "Likelihood",
                 M = mediators, Z = "late", Z.within = FALSE)
```

### Does time pressure moderate the expectation arm (Speed -> M)?

```{r bzmx-late}
knitr::kable(
  subset(mod_s$tidy, param == "BZ_MX")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Moderation of expectation paths by time pressure (BZ_MX)"
)
```

### Does it carry through to the indirect effect?

```{r bz-indirect-late}
knitr::kable(
  subset(mod_s$tidy, param == "BZ_MX * B1_YM")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Moderation of indirect effects by time pressure (BZ_MX x B1_YM)"
)
```

The largest BZ_MX is for `OnTime`: under time pressure, the expected
on-time gain from speeding is sharply amplified. The expected
meeting-quality gain (`IntQuality`) also strengthens. Other expectations
— including crash and ticket expectations — are not significantly
moderated by time pressure. People under time pressure don't expect
*different consequences* from speeding; they expect a *larger payoff* from
getting there on time.

The indirect-effect table propagates these shifts into the pull toward
speeding: under time pressure, the indirect path from Speed to Likelihood
strengthens through `OnTime` (the dominant route) and `IntQuality`. The
overall pull toward speeding rises with time pressure, not because the
costs feel smaller but because the on-time benefit feels bigger.

## Person-level moderation: self-reported fast-driver tendency

`SRFastDriver` is the [0, 1] rescaled response to *"I regularly drive far
over the speed limit when driving"* — item `SR_30` in `$traits`. As a
between-person trait, it uses `Z.within = FALSE`. Some respondents are
missing this item:

```{r prep-srfast}
sr <- speedingESJT$traits[, c("p","SR_30")]
sr <- sr[!is.na(sr$SR_30), ]
names(sr)[2] <- "SRFastDriver"
dat2 <- merge(dat, sr, by = "p")
cat("Persons with SRFastDriver:", length(unique(dat2$p)), "\n")

mod_p <- pathXMY(dat2, X = "Speed", Y = "Likelihood",
                 M = mediators, Z = "SRFastDriver", Z.within = FALSE)
```

### Does fast-driver tendency moderate the expectation arm?

```{r bz-mx-srfast}
knitr::kable(
  subset(mod_p$tidy, param == "BZ_MX")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Moderation of expectation paths by SRFastDriver (BZ_MX)"
)
```

### Does it carry through to the indirect effect?

```{r bz-indirect-srfast}
knitr::kable(
  subset(mod_p$tidy, param == "BZ_MX * B1_YM")[, c("mediator","est","se","z","pvalue")],
  digits = 3,
  caption = "Moderation of indirect effects by SRFastDriver (BZ_MX x B1_YM)"
)
```

Self-described fast drivers expect speeding to be less catastrophic
(smaller `Crash`, `Injured`, `Ticket`, and `MoneyCost` shifts), more
enjoyable (`FunDrive`), and substantially more role-appropriate
(`Appropriate` — by far the largest effect). These expectation differences
propagate into significantly stronger indirect pulls toward taking the
faster action through every meaningful pathway: `Appropriate` (the largest),
`FunDrive`, `Crash`, `Injured`, and `Ticket`. The contrast with the
time-pressure result is informative: time pressure shifts the *upside*
(OnTime, IntQuality); fast-driver tendency shifts the *downside* (the
expected costs of speeding feel smaller, and the action feels more
appropriate).

## Total moderation and its decomposition

The `BZ_MX × B1_YM` table above isolates one of three routes through
which a moderator `Z` shifts the `X → Y` link. The full identity is

$$\beta^Z_{YX}[1] \;=\; \beta^Z_{MX} \cdot \beta^1_{YM}
\;+\; \beta^1_{MX} \cdot \beta^Z_{YM}
\;+\; \beta^Z_{YX}[2]$$

where:

- **Term 1** (`BZ_MX × B1_YM`): `Z` shifts *what the action is expected
  to produce* — the **expectation** route highlighted above.
- **Term 2** (`B1_MX × BZ_YM`): `Z` shifts *how much the expected
  outcome matters* — the **valuation** route. Often overlooked.
- **Term 3** (`BZ_YX[2]`): residual direct `X × Z` moderation not
  routed through this mediator.

A moderator can produce a large total effect (`BZ_YX[1]`) through
*any* of these routes — and the standard `BZ_MX` table is silent
about Terms 2 and 3. `pathXMY_decompose()` fits both the no-mediator
model (giving `BZ_YX[1]`) and the full mediated model (giving the
three components) and tabulates them per mediator.

```{r decompose-srfast}
dec <- pathXMY_decompose(dat2, X = "Speed", Y = "Likelihood",
                         M = mediators, Z = "SRFastDriver",
                         Z.within = FALSE)
```

### The total moderation

```{r decompose-total}
knitr::kable(dec$total, digits = 3,
             caption = "Total moderation BZ_YX[1] (no mediator)")
```

Self-reported fast-driver tendency moderates the `Speed → Likelihood`
link by **+0.25** — a substantial effect: respondents one SD higher
on `SRFastDriver` increase their action-likelihood by an extra 0.25
units (on the [0, 1] outcome) when shifting from at-limit to
+10+ mph speeds.

### Decomposition per mediator

```{r decompose-table}
knitr::kable(dec$components, digits = 3,
             caption = "Decomposition of BZ_YX[1] through each single mediator")
```

Each mediator's three components plus residual sum to ~0.25 — that's
the algebraic identity at work. The `sum (1+2+3)` row should equal
`BZ_YX[1]` exactly in the linear case (small deviations come from
numerical precision).

**Where does the moderation route?** The largest single-mediator
expectation-route is through `Appropriate`: `BZ_MX × B1_YM = 0.073`.
Term 2 (valuation route) is negligible for every mediator here —
fast drivers don't differentially *weight* outcomes, they have
different *expectations*. And **most of `BZ_YX[1]` sits in the
residual `BZ_YX[2]` column** (≈ 0.18–0.24 for every mediator). That
isn't a failure of the decomposition — it's the signature of
moderation routed through *multiple mediators jointly*. With a
single mediator in the model, every other mediator's contribution
gets absorbed into the residual.

### Field-style diagrams

`plotPathXMY()` renders the X-M-Y structure in the funfield
visualization style:

- Nodes are colored by their **expected score when X is held at 1**
  — positive in blue, negative in red, gradient by magnitude. X is
  by definition full blue (score = 1); each mediator is colored by
  its `B1_MX` coefficient; Y is colored by the implied total effect
  at X=1.
- Y (Likelihood) is drawn as a left-facing triangle to signal that
  it's the focal action node.
- Edges are **dashed for negative**, **solid for positive**, with
  thickness scaled by path magnitude (`negDashed = TRUE`,
  `maximum = 0.8`). Edge labels show the full `b1 + bZ(Z)` form.

**Single-mediator triangle** for the dominant route through
`Appropriate`:

```{r plot-appropriate, fig.width = 7, fig.height = 5}
plotPathXMY(dec, mediator = "Appropriate",
            X_label = "Speed", Y_label = "Likelihood",
            Z_label = "SRFast",
            title = "Speed -> Appropriate -> Likelihood (Z = SRFastDriver)")
```

Reading the diagram: a one-unit increase in `Speed` lowers
`Appropriate` by 0.45 *on average* (dashed edge, negative), but
among fast drivers (high `SRFast`) that drop is dampened by +0.14
— they perceive faster driving as less of an appropriateness
violation. `Appropriate` strongly drives `Likelihood` (+0.53,
solid) and that valuation is *not* moderated by `SRFast`
(≈ +0.00). The residual `BZ_YX[2]` direct path (+0.18) reflects
the joint contribution of the seven other mediators not shown
here. The Y node renders red because the total predicted Y at
`Speed = 1` is negative — speeding produces a net-negative
expected likelihood when only `Appropriate` mediates.

**Full multi-mediator fan view** — pass no `mediator` argument and
all available mediators are arranged in a fan between X and Y:

```{r plot-fan, fig.width = 10, fig.height = 7}
plotPathXMY(dec,
            X_label = "Speed", Y_label = "Likelihood",
            Z_label = "SRFast",
            title = "Full X-M-Y field for speeding decisions")
```

The fan diagram shows the whole expectancy-value structure at a
glance. The mediator column reveals which expectations move with
Speed: `Ticket` (+0.55) and `Appropriate` (-0.45) are the strongest
in opposite directions; `FunDrive` and `IntQuality` are essentially
unaffected by Speed (near-white). The right column of edges shows
the valuation weights (`B1_YM`). Edge thickness in the fan
diagnoses where the indirect-effect mass concentrates.

### When does Term 2 carry the moderation?

A counter-example shows valuation routing in isolation. Consider
`DrivingSkill` (SR_07: *"I would say that I am a skilled driver"*)
as the moderator, with `Crash` as the focal mediator:

```{r decompose-skill}
tr <- speedingESJT$traits[, c("p","SR_07")]
names(tr)[2] <- "DrivingSkill"
tr <- tr[!is.na(tr$DrivingSkill), ]
dat_sk <- merge(dat, tr, by = "p")

dec_sk <- pathXMY_decompose(dat_sk, X = "Speed", Y = "Likelihood",
                            M = mediators, Z = "DrivingSkill",
                            Z.within = FALSE)
knitr::kable(
  subset(dec_sk$components, mediator == "Crash"),
  digits = 3,
  caption = "Crash decomposition under Z = DrivingSkill"
)
```

Here Term 1 (`BZ_MX × B1_YM = -0.002`) is essentially zero —
self-rated skilled drivers don't expect a *different* crash
likelihood from speeding. But Term 2 (`B1_MX × BZ_YM = -0.061`) is
significantly non-zero: skilled-driver respondents weight `Crash`
*more* heavily as a deterrent than unskilled-driver respondents
do. Without `pathXMY_decompose()`, this valuation moderation would
be invisible in the standard `BZ_MX` table.

```{r plot-skill, fig.width = 7, fig.height = 5}
plotPathXMY(dec_sk, mediator = "Crash",
            X_label = "Speed", Y_label = "Likelihood",
            Z_label = "Skill",
            title = "Speed -> Crash -> Likelihood (Z = DrivingSkill)")
```

## Scoring named trait composites

If you want the Big Five and other named scales from
Wood, Harms, & Cho (2023), use the recipes in
`speedingESJT$codebook$trait_scales`:

```{r scoring, eval = FALSE}
tr <- speedingESJT$traits
tr$Extra <- rowMeans(cbind(1 - tr$SR_15, tr$SR_16, tr$SR_17), na.rm = TRUE)
tr$Agree <- rowMeans(cbind(tr$SR_18, 1 - tr$SR_19, tr$SR_20), na.rm = TRUE)
tr$Consc <- rowMeans(cbind(1 - tr$SR_21, 1 - tr$SR_22, tr$SR_23), na.rm = TRUE)
tr$Neuro <- rowMeans(cbind(tr$SR_24, tr$SR_25, 1 - tr$SR_26), na.rm = TRUE)
tr$Open  <- rowMeans(cbind(tr$SR_27, 1 - tr$SR_28, tr$SR_29), na.rm = TRUE)
```

## Function reference

See `?pathXMY`, `?pathXMY_decompose`, `?plotPathXMY`, and
`?speedingESJT`.
