---
title: "Moderated mediation with `pathXMY()`: the speeding ESJT example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Moderated mediation with pathXMY(): the speeding ESJT example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette walks through fitting an expectancy-value (EV) moderated
mediation model to ESJT data using `pathXMY()`. We use the embedded
`speedingESJT` dataset (Wood, Harms, & Cho, 2023).

## The data

```{r data}
library(funfield)
data(speedingESJT)
str(speedingESJT[, 1:13])
```

302 participants each rated three speeds (at the limit, +5 mph, +10 mph) on
their expected outcomes (`Crash`, `Injured`, `Ticket`, `MoneyCost`, `OnTime`,
`IntQuality`, `FunDrive`, `Appropriate`) and on the likelihood `L` that they
would take that action.

## Within-person deviation

`pathXMY()` fits a single-level path model with cluster-robust standard
errors. To get correct within-person coefficients, deviate the Level 1
variables before fitting:

```{r deviate}
L1 <- c("Speed","Crash","Injured","Ticket","MoneyCost","OnTime",
        "IntQuality","FunDrive","Appropriate","L")
dev <- speedingESJT
for (v in L1) {
  dev[[v]] <- ave(dev[[v]], dev$p, FUN = function(x) x - mean(x))
}
```

`Z` (the trait moderator) is left as-is — `pathXMY()` will z-standardize it
internally.

## Normative model (no moderator)

How does each potential outcome mediate the Speed -> Likelihood relationship,
averaging across people?

```{r normative}
mediators <- c("Crash","Injured","Ticket","MoneyCost","OnTime",
               "IntQuality","FunDrive","Appropriate")
norm <- pathXMY(dev, X = "Speed", Y = "L", M = mediators)
norm$tidy
```

Each mediator gets six rows: `B1_MX` (expectation: does Speed predict M?),
`B1_YX` (direct effect of Speed on L not through this M), `B1_YM`
(valuation: does M predict L?), and `B1_MX * B1_YM` (the indirect product).

## Moderated model: high vs. low SRFastDriver

Does self-reported fast-driver trait moderate any of those paths?

```{r moderated}
mod <- pathXMY(dev, X = "Speed", Y = "L", M = mediators,
               Z = "SRFastDriver")

## The Z-moderation of the expectation arm (Speed -> M):
subset(mod$tidy, param == "BZ_MX")[, c("mediator","est","se","z","pvalue")]

## The Z-moderation of the indirect effect (via expectation):
subset(mod$tidy, param == "BZ_MX * B1_YM")[, c("mediator","est","se","z","pvalue")]

## The Z-moderation of the valuation arm (M -> L):
subset(mod$tidy, param == "BZ_YM")[, c("mediator","est","se","z","pvalue")]
```

`BZ_MX` captures how the Speed -> M expectation differs across trait levels
(self-reported fast drivers expect speeding to be less catastrophic and more
enjoyable). `BZ_MX * B1_YM` is the corresponding Z-moderation of the
indirect effect.

## Bootstrap option for small samples

With many clusters (here 302), delta-method CIs and cluster bootstrap CIs
agree closely. For studies with fewer than 50 clusters, use `se = "boot"`:

```{r boot, eval = FALSE}
pathXMY(dev, X = "Speed", Y = "L", M = "Crash",
        Z = "SRFastDriver", se = "boot", nboot = 1000)
```

## Function reference

See `?pathXMY` and `?speedingESJT`.
